---
title: "Regression 2"
author: "Aliaksandr Hubin"
date: "February 17 2026"
output:
  ioslides_presentation: default
  slidy_presentation: default
editor_options:
  markdown:
    wrap: 72
---

## Topics for this lecture

-   Predictions (with uncertainty)
-   Significance of regression, the F-test
-   Interaction effects
-   Model checking
-   Polynomial regression
-   Transformations of the response
-   Variable selection

## The body data

Let's again look at the body data and the model: $$ 
y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 x_{3i}+\epsilon_i
$$ with the usual assumption of independent errors distributed as
$\epsilon_i \sim N(0, \sigma^2)$. Here $y_i$ is the body weight of
person $i$, $x_{1i}$ the height, $x_{2i}$ is the circumference and
$x_{3i}$ is the age of the same person.

```{r, echo=FALSE}
library(BIAS.data)
```

## Our model with 3 predictors

```{r, echo=FALSE}
LinearModel.4 <- lm(Weight ~ Height + Circumference + Age, data=bodydata, 
  subset=1:20)
sum.model4 <- summary(LinearModel.4)
print(sum.model4)
```

## Prediction

A fitted model can be used for prediction of new observations

According to model 4, what is the predicted Weight of some person with
these values?

```{r, echo=FALSE}
newObs <- data.frame(Height=171, Circumference=76, Age=30)
print(newObs)
```

$\widehat{y} = -66.579 + 0.4768\cdot171 + 0.7769\cdot76 - 0.2094\cdot30$

$= 67.72$

```{r, echo=TRUE}
newobs <- data.frame('Height'=171, 'Circumference'=76, 'Age'=30)
predict.lm(LinearModel.4, newdata=newobs)
```

## Confidence in predictions

**Confidence interval for**
$E(Y|x_1, \ldots, x_p) \in \widehat{\mu}(\mathbf{x}_{\text{new}}) \pm t_{\alpha/2,\,n-p-1} \cdot \text{SE}(\widehat{\mu}(\mathbf{x}_{\text{new}}))$
where $\text{SE}(\widehat{\mu}(\mathbf{x}_{\text{new}}))$ accounts for
uncertainty in estimation.

```{r, echo=TRUE}
predict.lm(LinearModel.4, newobs, level=0.95, interval=c("confidence"))
```

**Prediction interval for**
$Y|x_1, \ldots, x_p \in \widehat{\mu}(\mathbf{x}_{\text{new}}) \pm
t_{\alpha/2,\,n-p-1}\,\text{SE}_{\text{pred}}$

```{r, echo=TRUE}
predict.lm(LinearModel.4, newobs, level=0.95, interval=c("prediction"))
```

## Predicted vs observed - plot

```{r, echo=FALSE, out.width=700, out.height=400}
pred <- predict(LinearModel.4, newdata=bodydata[-c(1:20),], 
  interval="prediction", level=0.95)
conf <- predict.lm(LinearModel.4, newdata=bodydata[-c(1:20),], 
           level=0.95, interval=c("confidence"))

newWeight <- bodydata[-c(1:20),1]

ord <- order(newWeight)
x   <- newWeight[ord]

## central fits
fit_mean <- pred[ord, 1]         # or your mean-prediction object
fit_mean_s <- predict(loess(fit_mean ~ x, span = 0.5))

## mean CI
lwr_ci <- conf[ord, 2]
upr_ci <- conf[ord, 3]
lwr_ci_s <- predict(loess(lwr_ci ~ x, span = 0.5))
upr_ci_s <- predict(loess(upr_ci ~ x, span = 0.5))

## predictive PI
lwr_pi <- pred[ord, 2]
upr_pi <- pred[ord, 3]
lwr_pi_s <- predict(loess(lwr_pi ~ x, span = 0.5))
upr_pi_s <- predict(loess(upr_pi ~ x, span = 0.5))

plot(x, fit_mean, xlab = "Observed", ylab = "Predicted")
#curve(1 * x, add = TRUE, col = 2, lwd = 2)

## mean fit
#lines(x, fit_mean_s, col = "red", lwd = 2)

## confidence band for mean
lines(x, lwr_ci_s, col = "blue",  lwd = 1.5, lty = 2)
lines(x, upr_ci_s, col = "blue",  lwd = 1.5, lty = 2)

## prediction band (wider)
lines(x, lwr_pi_s, col = "darkgreen", lwd = 1.5, lty = 3)
lines(x, upr_pi_s, col = "darkgreen", lwd = 1.5, lty = 3)



```

## Check predictions' errors

```{r, echo=FALSE}

head(pred,12)

print(paste0(" RMSE  observed: ", round(sqrt(mean((predict(LinearModel.4,  newdata=bodydata[c(1:20),])-bodydata[c(1:20),1])^2)),4)))
print(paste0(" RMSE  predictions: ", round(sqrt(mean((pred[,1]-bodydata[-c(1:20),1])^2)),4)))

```

## Checking the prediction intervals

How many observations fell within the prediction intervals?

```{r, echo=TRUE}
largerthanlower <- (newWeight > pred[,2])
smallerthanupper <- (newWeight < pred[,3])
inInterval <- which(largerthanlower & smallerthanupper)
print(length(inInterval))
```

Proportion of observations within prediction intervals:

```{r, echo=TRUE}
length(inInterval)/length(newWeight)
```

What should we expect?

## SSE, SST, SSR again

-   Total Sum of Squares: $SST = \sum_{i=1}^n (y_i - \bar{y})^2$

-   Error Sum of Squares:
    $SSE = \sum_{i=1}^n (y_i - \widehat{\mu}(x_i))^2$

-   Regression Sum of Squares:
    $SSR = \sum_{i=1}^n (\widehat{\mu}(x_i) - \bar{y})^2$

    It can be shown:

$SST = SSE + SSR \quad \Rightarrow\quad R^2 = 1 - \frac{SSE}{SST} = \frac{SSR}{SST}$

## Significance of regression?

We have seen how we can test the significance of single predictor terms
with T-tests in the previous lesson.

We may also test the overall significance of the regression model.

Recall from earlier that $R^2 = 1-\frac{SSE}{SST} = \frac{SSR}{SST}$ is
a measure of model fit. If the model can explain much of the observed
variability in $y$, then SSR is close to SST and big compared to SSE.

Can be shown that a convenient test observator, typically named $F$ (for
Fisher) is:

$$F = \frac{SSR/p}{SSE/(n-p-1)}$$

## The F-test for overall significance of regression

Hypothesis:

$H_0: \beta_1 = \beta_2 = \cdots = \beta_p = 0$ vs $H_1:$ at least one
$\beta_j \neq 0$.

Reject the null hypothesis at test level $\alpha$ if $F$ is large, and
more specifically if

$F > F_{\alpha, p, n-p-1}$

Let's see how this turns out for the body data model.

## The F-test illustration

## The body data

```{r, echo=TRUE, eval=FALSE}
body1 <- lm(Weight ~  Height + Circumference + Age, data=bodydata, subset = 1:20)
Atab <- anova(body1)
```

```{r, echo=FALSE}
library(BIAS.data)
data(bodydata)
body1 <- lm(Weight ~  Height + Circumference + Age, data=bodydata, subset = 1:20)
Atab <- anova(body1)
print(Atab)
SST <- sum(Atab[1:4,2])
SSR <- SST - Atab[4,2]
CSST <- cumsum(Atab[1:4,2])
cat('', '\n')
cat(' SST = ', SST, ', SSR = ', SSR, ';\n Cummulative SST', CSST, ';\n F = ', (SSR/3)/(Atab[4,2]/Atab[4,1]), '\n')

```

## The body data - Compare with previous

```{r, echo=TRUE, eval=FALSE}
body1 <- lm(Weight ~  Height + Circumference + Age, data=bodydata, subset = 1:20)
Atab <- anova(body1)
```

```{r, echo=FALSE}
library(BIAS.data)
data(bodydata)
body1 <- lm(Weight ~ Circumference + Height + Age, data=bodydata, subset = 1:20)
Atab <- anova(body1)
print(Atab)
SST <- sum(Atab[1:4,2])
SSR <- SST - Atab[4,2]
cat(' ', '\n')
cat('SST = ', SST, ', SSR = ', SSR, ', F = ', (SSR/3)/(Atab[4,2]/Atab[4,1]), '\n')

```

## Interaction effects

Interaction effects occur when the effect of one variable on the
response variable depends on the level/value of another variable.

Consider the model: $$
y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 x_{3i} + \beta_4 x_{2i}x_{3i} + \epsilon_i
$$

here we have expanded the model with an interaction term between $x_2$
and $x_3$ (between circumference and age). How should we interpret this
model?

## How to fit the model in RStudio

Now fit on all the data we have

```{r, echo=TRUE, eval=FALSE}
data(bodydata)
body2 <- lm(Weight ~ Height + Circumference + Age + 
              Circumference:Age, data=bodydata)
summary(body2)
```

## The output with all data

```{r, echo=FALSE}
data(bodydata)
body2 <- lm(Weight ~ Height + Circumference + Age + Circumference:Age, data=bodydata)
summary(body2)
```

## Model checking

In standard linear regression we typically assume:

-   Linearity
-   Constant variance
-   Independent observations
-   Normality

We should check that these assumptions are OK. Typically we do this by
examining the residuals

## Residual plots

```{r, echo=TRUE}
par(mfrow=c(2,2), mar=c(4.5,4,1.5,1.5))
plot(body2, ask=FALSE)
```

## Draw Q-Q plot and std residuals

## Another example - Bears

```{r, echo=FALSE}
data(bears)
bears1 <- lm(Weight ~ Length, data=bears)
```

```{r, echo=FALSE}
plot(Weight ~ Length, data=bears, pch=20, cex=2, ylim=c(0,250))
abline(bears1, col=2, lwd=2)
```

## Summary

```{r, echo=FALSE}
summary(bears1)
```

## Residual plots

```{r, echo=TRUE}
par(mfrow=c(2,2), mar=c(4.5,4,1.5,1.5))
plot(bears1, ask=FALSE)
```

## Outliers and influence statistics

-   Leverage
-   Standardized residuals
-   Deleted residuals (PRESS residuals) (predict the deleted leave out
    one point)
-   Cook's distance (combines leverage and standardized residuals) \> 1,
    can be dangerous

## Explain leverage

## Polynomial regression

Non-linear relationships may be modelled by polynomial regression.

Bear example

-   $y=$ Weight
-   $x_1$ = Length
-   $x_1^2$ = Length$^2$

A multiple LINEAR model (linear in the $\beta$'s)

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_1^2 + \epsilon$$

## Fitting the model in R

```{r, echo=TRUE}
data(bears)
bears2 <- lm(Weight ~ Length + I(Length^2), data=bears)
summary(bears2)
```

## A fitted line plot

```{r, echo=FALSE}
plot(Weight ~ Length, data=bears)
curve(291.62 -5.29*x + 0.026*x^2, add=TRUE, lwd=2, col="red")
```

## Careful! Over-fitting!

```{r, echo=FALSE}
data(bears)
bears3 <- lm(Weight ~ Length + I(Length^2) + I(Length^3) + I(Length^4) + I(Length^5), data=bears)
pred <- predict(bears3, newdata=data.frame(Length=seq(90, 190, 1)))
plot(Weight ~ Length, data=bears, main=bears3$call, cex.main=0.7)
lines(seq(90, 190, 1), pred, col="red", lwd=2)
```

## Careful! Over-fitting!

```{r, echo=FALSE}
data(bears)
bears3 <- lm(Weight ~ Length + I(Length^2) + I(Length^3) + I(Length^4) + I(Length^5) + I(Length^6) + I(Length^7) + I(Length^8) + I(Length^9) + I(Length^10), data=bears)
pred <- predict(bears3, newdata=data.frame(Length=seq(90, 190, 1)))
plot(Weight ~ Length, data=bears, main=bears3$call, cex.main=0.7)
lines(seq(90, 190, 1), pred, col="red", lwd=2)
```

## Careful! Over-fitting!

```{r, echo=FALSE}
data(bears)
bears3 <- lm(Weight ~ + poly(Length,15), data=bears)
pred <- predict(bears3, newdata=data.frame(Length=seq(90, 190, 1)))
plot(Weight ~ Length, data=bears, main=bears3$call, cex.main=0.7)
lines(seq(90, 190, 1), pred, col="red", lwd=2)
```

## Transformations

Deviations from model assumptions may be fixed by transforming $y$
and/or $x$.

Some transformations which may improve linearity and stabilize variance
are

-   $\widetilde{y} = log(y)$
-   $\widetilde{x} = log(x)$
-   $\widetilde{y} = \sqrt{y}$

Then use $\widetilde{y}$ and/or $\widetilde{x}$ instead of $y$ and $x$
to fit the model

## The log(Weight) response model

```{r, echo=TRUE}
bears4 <- lm(log(Weight) ~ Length, data=bears)
summary(bears4)
```

## Explain the transformation and back-transformation

## Residual plots

```{r, echo=TRUE}
par(mfrow=c(2,2), mar=c(4.5,4,1.5,1.5))
plot(bears4, ask=FALSE)
```

## A fitted line plot

```{r, echo=FALSE}
plot(Weight ~ Length, data=bears)
curve(exp(0.71 + 0.025*x), add=TRUE, lwd=2, col="red")
```

## Variable selection

Variable selection is the process of selecting a subset of variables
that are most relevant for the response in some way. Various approaches.
Some are based on hypothesis testing, some on criteria (like Akaike's
Information Criteria - AIC, Bayesian information criterion - BIC,
adjuseted $R^2_{adj}$, etc)

Sometimes combined with stepwise approaches like:

-   Forward selection
-   Backward selection
-   Forward/Backward
-   Backward/Forward
-   All subsets selection

```{r, echo=TRUE}
library(mixlm)
```

## Stepwise methods - Backward selection

```{r, echo=TRUE, eval=TRUE}
body1 <- lm(Weight ~ Height + Circumference + Age, data=bodydata, subset=1:50)
backward(body1, alpha=0.05)
```

## Stepwise methods - Forward selection

```{r, echo=TRUE, eval=TRUE}
body1 <- lm(Weight ~ Height + Circumference + Age, data=bodydata, subset=1:50)
fs <- forward(body1, alpha=0.05)
```

## Best subsets method

If the number of candidate variables, $p$, is limited, e.g. $\leq20$, we
may compare all possible models

There are $2^p$ possible models

Compare all model fits by AIC, BIC, $R^2_{adj}$

Choose model based on:

-   Best-fit - simplicity trade-off

## Best subsets method - Example

```{r, echo=TRUE, eval=TRUE}
body1 <- lm(Weight ~ Height + Circumference + Age, data=bodydata, subset=1:50)
bs <- best.subsets(body1)
```
