---
title: "ANOVA 1"
author: "Aliaksandr Hubin"
date: "Feb 24 2025"
output: 
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
  # output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

## Topics for this lecture

-   Introductory example

-   "ANOVA"

-   The F-test

-   Estimation and parametrization

-   Post hoc contrasts and pairwise tests


## Objectives of lecture

-   *Basic framework*: The response variable is continuous as for
    regression but the predictor variables are *factors* like
    fertilizer, soil type, country, ...
-   *Basic idea*: The statistical analysis is based on studying the
    variability in data. Hence the name ANOVA:
    -   ANalysis Of VAriance.
-   Objectives as for regression: predict and explain

## Example Yields

Yield from a crop experiment on $k=3$ different soil types. There are
$n_i=10$ random replicates per soil type. A total of $n=k\cdot n_i = 30$
observations altogether.

```{r, echo=FALSE, message=FALSE}
library(BIAS.data)
library(mixlm)
data(yields)
attach(yields)
plot(yield ~ soil, data=yields, col="green",main = "Are the expected yields different?")
```

## Linear Regression Formulation

Model with dummy variables for soil type:

$$y_j = \beta_0 + \beta_1 x_{1j} + \beta_2  x_{2j} + \beta_3  x_{3j}  + \epsilon_j,\epsilon_j \sim N(0, \sigma^2) $$

where:

-   $x_{1j} = 1$ if soil is clay, else 0
-   $x_{2j} = 1$ if soil is loam, else 0
-   $x_{3j} = 1$ if soil is sand, else 0
-   $\beta_0 = \mu$ = intercept (grand mean)
-   $\beta_0 + \beta_1 = \mu_1$ = mean of yield for clay
-   $\beta_0 + \beta_2= \mu_2$ = mean of yield for loam
-   $\beta_0 + \beta_3 = \mu_3$ = mean of yield for sand


## Linear Regression Reformulate

Model with dummy variables for soil type:

$$y_j = \beta_0 + \beta_2  x_{2j} + \beta_3  x_{3j}  + \epsilon_j,\epsilon_j \sim N(0, \sigma^2) $$

where:

-   $x_{2j} = 1$ if soil is loam, else 0
-   $x_{3j} = 1$ if soil is sand, else 0 and hence $x_{1j} = 1 - x_{2j} - x_{3j}$
-   $\beta_0 = \mu_1$ = intercept (mean of yield for clay)
-   $\beta_0 + \beta_2= \mu_2$ = mean of yield for loam
-   $\beta_0 + \beta_3 = \mu_3$ = mean of yield for sand 
-   same model but with $p=2$ not $p=3$.

## ANOVA

Let us use two indices instead of dummy variables

-   $i$ - indexes the group (soil type): $i \in \{1, 2, 3\}$
-   $j$ - indexes observation within group: $j \in \{1, \ldots, n_i\}$

Let $y_{ij}$ be the observed yield for observation number
$j \in \{1,\ldots, n_i\}$ soil type $i \in \{1,\ldots, k\}$.

The model (One-way ANOVA): $$y_{ij} = \mu_{i} + \epsilon_{ij}$$

where $\epsilon_{ij} \sim N(0, \sigma^2)$ and $\mu_i$ is the expected
yield for soil type $i$.

## Splitting up total variability

SST=
$\sum_{i=1}^{k}\sum_{j=1}^{n_i}(y_{ij}-\bar{y}_{..})^2, \quad \bar{y}_{..}\text{ is grand mean}$

For the yield data:

```{r, echo=FALSE}
sum((yield-mean(yield))^2)
```

SSA = $\sum_{i=1}^{k}\sum_{j=1}^{n_i}(\bar{y}_{i.}-\bar{y}_{..})^2 =
n\sum_{i=1}^{k}(\bar{y}_{i.}-\bar{y}_{..})^2,  \quad \bar{y}_{i.}\text{i-th group mean}$

For the yield data:

```{r, echo=FALSE}
soilmeans <- by(yield, soil, mean)
10*sum((soilmeans - mean(yield))^2)
```

-   SSE = SST - SSA =
    $\sum_{i=1}^{k}\sum_{j=1}^{n_i}(y_{ij}-\bar{y}_{i.})^2$ = 414.7 - 99.2
    = 315.5

## Between and within group variances:

Dividing Sums of Squares by their respective degrees of freedom gives:

-   The between soil type (group) variance:
    $MSA = SSA/(k-1) = 99.2/2 = 49.6$
-   The within soil type (group) variance:
    $MSE = SSE/(n-k) = 315.5/27 = 11.685$

where $n = n_1 + n_2 + \cdots n_k$ is the total number of observations.

**Why does it make sense to compare these?**

## Testing the effect of soil

If there is a soil effect on yield, the expected yields will differ in
one way or the other.

Hypotheses:

$$H_0: \mu_1 = \mu_2 =\mu_3 $$ vs $$H_1: \mu_i \neq \mu_{i'}$$ for at
least two different soil types $i$ and $i'$.

Reject $H_0$ if $MSA >> MSE$

## Test observator, F

If $H_0$ is true, then

$$F = \frac{MSA}{MSE} \sim F(\nu_1 = k-1 = 3 -1, \nu_2 = n - k = 30 - 3)$$

```{r, echo=FALSE}
curve(df(x,2, 27), xlim=c(0,7), lwd=2, col=2, ylab="Fisher(2,27)", xlab="F")
abline(h=0, v=c(0, 4.24))
text(5.5, 0.2, labels="F = 49.6/11.7 = 4.24")
```

## Rejection region

Reject $H_0$ at test level $\alpha$ if
$F_{obs}> F_{\alpha, \nu_1, \nu_2}$, where $\nu_1 = k-1$ are the degrees
of freedom for MSA and $\nu_2 = n -k$ are the degrees of freedom for
MSE.

For example, assume a test level of 5%

From a Fisher table we find $F_{0.05, 2, 27}$ = 3.35

Alternatively, reject if p-value is smaller than $\alpha$.

**Conclusion?**

## Estimation of group effects and error variance

The inutitive estimators for $\mu_i$ are
$\widehat{\mu}_i = \bar{y}_{i.}$

These are also the least squares estimators (for balanced data), which
are unbiased and have minimum variance among all unbiased estimators.

As in regression analysis we have $\widehat{\sigma}^2 = MSE$.

## Splitting up $\mu_i$ - i.e. reparametrise

The means may be split up further. Assume

$$\mu_i = \mu + \alpha_i$$

But, all parameters can not be estimated without some extra restriction
(choice of parametrization).

-   Parametrization 1: Assume $\sum_{i=1}^k \alpha_i = 0$

A "sum to zero" parametrization ("contr.sum" in R")

-   Parametrization 2: Assume $\alpha_1 = 0$

A "reference level" or "control level" parametrization
("contr.treatment" in R")

## Choice of arametrization in RStudio

Parametrization 1 (Using the lm-function in the mixlm-package)

```{r, echo=TRUE, eval=FALSE}
mod1 <- lm(yield~soil, data=yields, contrasts = "contr.sum")
```

Parametrization 2 (Using the lm-function in the mixlm-package

```{r, echo=TRUE, eval=FALSE}
mod1 <- lm(yield~soil, data=yields, contrasts = "contr.treatment")
```

## Example: Parametrization 1

```{r, echo=FALSE}
mod1 <- lm(yield~soil, data=yields, contrasts = "contr.sum")
round(summary(mod1)$coef,4)
```

Compare with group and total means:

```{r}
means <- c(soilmeans,mean(soilmeans))
names(means) <- c("clay", "loam", "sand", "total-mean")
print(means)
```

## Example: Parametrization 2

```{r, echo=FALSE}
mod1 <- lm(yield~soil, data=yields, contrasts = "contr.treatment")
round(summary(mod1)$coef,4)
```

Compare with group and total means:

```{r, echo=FALSE}
means <- c(soilmeans,mean(soilmeans))
names(means) <- c("clay", "loam", "sand", "total-mean")
print(means)
```

## ANOVA table of Sum of squares

This table does not depend on parametrization

```{r, echo=FALSE}
anova(mod1)
```

## Residual analysis

We should also for ANOVA-models check the assumptions by residual
analysis

```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(mod1, which=c(1,2))
```

## Pair-wise contrasts

If a factor is significant and has more than two levels: Where are the
differences?

May perform a "post hoc" test with contrasts.

A contrasts is a linear function of the factor effects.

$$\theta = c_1 \mu_1 + c_2 \mu_2 + c_3 \mu_3$$

where $\sum_{i=1}^k c_i= 0$

For pair-wise contrasts, only two $c_i$'s differ from 0.

## Example 1 (With parametrization 1)

```{r, echo=FALSE}
options(contrasts=c("contr.sum","contr.poly")) 
mod1 <- stats::lm(yield~soil, data=yields)
```

Comparing soil type "loam" (level 1) with "clay" (level 2):

$\theta_1 = 1\mu_1 + (-1)\mu_2 + 0 \mu_3$

$= 1(\mu + \alpha_1) - 1(\mu + \alpha_2) = \alpha_1 -\alpha_2$

Estimate:
$\widehat{\theta}_1 = \widehat{\alpha}_1 -\widehat{\alpha}_2 = -0.4 - 2.4 = -2.8$

```{r, echo=TRUE}
gmodels::fit.contrast(model=mod1, varname='soil', df=TRUE, coeff=c(1, -1, 0), conf.int=0.95)
```

## Example 2: "loam" vs "sand"

Define required pairwise contrast:

$\theta_2 = 0\mu_1 + 1\mu_2 - 1\mu_3$
$= 1(\mu + \alpha_2) - 1(\mu + \alpha_3) = \alpha_2 - \alpha_3$

Estimation

$\widehat{\alpha}_3 = -(-0.4 + 2.4) = -2$
$\widehat{\theta}_2 = \widehat{\alpha}_2 - \widehat{\alpha}_3 = 2.4 - (-2) = 4.4$

```{r, echo=FALSE}
gmodels::fit.contrast(model=mod1, varname='soil', df=TRUE, coeff=c(0, 1, -1), conf.int=0.95)
```

## Multiple testing - Tukey test

Type 1: $\alpha$ = P(Rejecting $H_0$ \| $H_0$ is TRUE)

If there are many levels of a factor, a Tukey test is preferable.

Controls overall error rate (family-wise error) of all pairwise tests.

Use 'simple.glht'-function in the mixlm-package.

```{r, echo=TRUE, eval=FALSE}
ptests <- mixlm::simple.glht(mod1, "soil", corr = c("Tukey"),level = 0.95)
print(ptests)
plot(ptests)
```

## R Commander output - Tukey

```{r, echo=FALSE}
ptests <- mixlm::simple.glht(mod1, "soil", corr = c("Tukey"),level = 0.95)
print(ptests)
```

## Pairwise confidence intervals plot

```{r, echo=FALSE}
par(mar=c(5,6,4,1))
plot(ptests)
```

## The ANOVA applet

The ANOVA "idea": Studying means by splitting variability

[ANOVA shiny app](https://solve.shinyapps.io/ANOVA/)


## Topics for next lecture

-   ANOVA with several factors
-   The factorial applet
-   Effect plots
-   Estimation and hypothesis testing
-   Tukey test and confidence intervals
-   ANCOVA - <Example:Regression> with indicator variables

## Announcements

- Elias is travelling, online session this Thursday 

-[What do we do with mondays](https://www.mentimeter.com/app/presentation/alvmjssjcyvd8zdzrwxw3tqapgghqi35/present?question=3v4by4qckmhx)

- Assingnment 1 will be published as a quizz next Tuesday, deadline March 10. 70% needed to pass.

- Practise quiz on canvas!

