---
title: "Regression 1"
author: "Aliaksandr Hubin"
date: "February 10 2025"
output:
  ioslides_presentation: default
  slidy_presentation: default
  beamer_presentation: default
 # output: word_document
---

## The body data

A motivating example - The body data

```{r, echo=FALSE, message=FALSE}
library(BIAS.data)
library(car)
pairs(bodydata, panel=panel.smooth)
```

## The Weight versus Height

Is there a linear relationship between Weight and Height?

```{r, echo=FALSE}
plot(Weight ~ Height, data=bodydata)
```

If so, what is the "best" fitting line?

## A graphical presentation of a simple linear model

Let's use a web applet to discuss the "best" fitting line: <http://solve.shinyapps.io/LeastSquaresApp>

## The main idea

A continuous response variable, $y$, is assumed to be linearly dependent on one or several predictor variables, $x_1, x_2, \ldots, x_k$?

A simple regression model (one predictor) $$ y = \beta_0 + \beta_1 x + \epsilon $$

A multiple regression model (multiple predictors) $$ y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots+ \beta_k x_k +\epsilon $$

where we typically assume $$ \epsilon \sim N(0, \sigma^2) $$

## Estimation of regression coefficients

Unknown regression coefficients of a simple regression model are $\beta_0$ and $\beta_1$

Estimation of parameters is equivalent to finding the "best fitting" model.

Different estimation methods $\Leftrightarrow$ Different definitions of "best".

-   Least squares estimator : The estimates minimize $\sum (y_i - (\widehat{\beta}_0 + \widehat{\beta}_1 x_i))^2$
-   Equivalent to the maximum likelihood estimator: The estimates make the observed data "most likely"

## Maximum likelihood formulation

## Derivations for maxima

## Estimation of unexplained variance

As soon as we have defined the "best" fitting model, we have also defined all distances between the observations and the estimated model.

These distances are the residuals: $$d_i = y_i - (\widehat{\beta}_0 + \widehat{\beta}_1 x_i)$$ for $i=1,\ldots,n$

An unbiased estimator for the error variance $\sigma^2$ depends on these distances through $$ \widehat{\sigma}^2 = \sum d_i^2/(n-p) = \frac{SSE}{(n-p)} = MSE$$

where $p$ is the number of estimated regression coeffiscients.

## Example: Regression in R (Model 1)

We use samples 1 to 20 of the bodydata as example data (training set)

$$ y_i = \beta_0 +\beta_1 x_{1i} + \epsilon_i $$ where $y$ is Weight, $x_1$ is Height and $i=1,\ldots, 20$.

## Example of model fit in R Studio

```{r, echo=TRUE, eval=FALSE}
LinearModel.1 <- lm(Weight ~ Height, data=bodydata, subset=1:20)
sum.model1 <- summary(LinearModel.1)
print(sum.model1)
```

## Output

```{r, echo=FALSE}
LinearModel.1 <- lm(Weight ~ Height, data=bodydata, subset=1:20)
sum.model1 <- summary(LinearModel.1)
print(sum.model1)
```

## The fitted line plot

```{r, echo=TRUE, eval=TRUE}
plot(Weight~Height, data=bodydata, subset=1:20, pch=20, cex=2)
abline(LinearModel.1, lwd=2, col="red")
```

## Weight versus Circumference (Model 2)

Now let:

$$ y_i = \beta_0 +\beta_1 x_{2i} + \epsilon_i $$

where $y$ still is Weight, $x_2$ is Circumference and $i=1,\ldots, 20$.

## R Commander output

```{r, echo=FALSE}
LinearModel.2 <- lm(Weight ~ Circumference, data=bodydata, subset=1:20)
sum.model2 <- summary(LinearModel.2)
print(sum.model2)
```

How can we claim that this model gives a better fit than model1?

## A measure of goodness of fit, the $R^2$

The unconditional (sample) variance of Weight is: $$\sum (y_i - \bar{y})^2/(n-1) = \frac{SST}{(n-1)}$$ Hence, $SST$ (in some literature $SSY$) is a mesure of total variability in y, the response

Further, $SSE$ is a measure of unexplained variability in y

So, a measure of explained variability is $$R^2 = \frac{SST-SSE}{SST} = \frac{SSR}{SST}$$

## Multiple linear regression (Model 3)

$$y_i = \beta_0 + \beta_1 x_{1i} + \beta_2x_{2i} + \epsilon_i$$

where $y$ is Weight, $x_1$ is Height $x_2$ is Circumference and $i=1,\ldots, 20$.

## R Studio output (Model 3)

```{r, echo=FALSE}
LinearModel.3 <- lm(Weight ~ Height + Circumference, data=bodydata, 
  subset=1:20)
sum.model3 <- summary(LinearModel.3)
print(sum.model3)
```

## Three predictors...(Model 4)

```{r, echo=FALSE}
LinearModel.4 <- lm(Weight ~ Height + Circumference + Age, data=bodydata, 
  subset=1:20)
sum.model4 <- summary(LinearModel.4)
print(sum.model4)
```

## $R^2$-adjusted

$$R^2 = 1 - \frac{SSE}{SST} \text{, but the Problem: $R^2$ always increases when adding predictors}$$

$$R^2_{\text{adj}} = 1 - \frac{\hat\sigma^2}{\widehat{Var}(y)} = 1 - \frac{SSE/(n-p)}{SST/(n-1)}$$ The Solution: $R^2_{\text{adjusted}}$ penalizes adding parameters

```{r, echo=FALSE, out.width=400}
r2 <- c(sum.model1$r.squared, sum.model2$r.squared,
  sum.model3$r.squared, sum.model4$r.squared)
r2adj <- c(sum.model1$adj.r.squared, sum.model2$adj.r.squared,
  sum.model3$adj.r.squared, sum.model4$adj.r.squared)
par(mar=c(4,4,0,0))
plot(1:4, r2adj, type="b", lwd=2, col=2, ylab="R2-adjusted", 
     xlab="Model complexity", xlim=c(1,7), ylim=c(0.3, 1))
points(1:4, r2, type="b", lwd=2, col=4)
```

## Hypothesis Test for Regression Coefficients

We may test the **conditional significance** of any predictor using a t-test

**Hypotheses:**

-   $H_0: \beta_j = 0$ (predictor has **no effect** on response)

-   $H_1: \beta_j \neq 0$ (predictor **does have an effect**)

**Test statistic:** $$T_j = \frac{\widehat{\beta}_j-0}{SE(\widehat{\beta}_j)} \sim t_{n-p}$$

## Visual explanation

## Decision rule

The **p-value** is the probability of observing a test statistic as extreme or more extreme than the observed value, **assuming** $H_0$ is true

$$\text{p-value} = P(|T| \geq |T_{\text{obs}}|) = 2 \times P(T > |T_{\text{obs}}|)$$

-   **p-value is small** (e.g., 0.002): It's **very unlikely** to observe such an extreme
-   T if $H_0$ were true → Strong evidence against $H_0$ → **Reject** $H_0$

**At significance level of 0.05:**

-   Reject $H_0$ if $|T_j| > t_{\alpha/2, n-p}$ (i.e., p-value $< 0.05$)
-   Fail to reject $H_0$ if $|T_j| \leq t_{\alpha/2, n-p}$ (i.e., p-value $\geq 0.05$)

## Example: Model 4 Significance tests

```{r, echo=FALSE}
sum.model4$coef
```

**Key insight:**

-   Age is not significant (p = 0.260 \>\> 0.05)

-   Explains why adding Age slightly reduces $R^2_{\text{adj}}$ - Model 3 (without Age) is **better choice** than Model 4

-   Other predictors are significant

## Confidence interval for regression coefficients

A $(1-\alpha)\cdot 100 \%$ for $\beta_j$ is mathematically given by: $$ \left[ \widehat{\beta}_j\pm t_{\alpha/2, (n-p)} SE(\widehat{\beta}_j) \right]$$

In RStudio we can get these (ex. for model 4) by

```{r, echo=TRUE, eval=TRUE}
confint(LinearModel.4, level=0.95)
```

## Prediction

A fitted model can be used for prediction of new observations

According to model 4, what is the predicted Weight of some person with these values?

```{r, echo=FALSE}
newObs <- data.frame(Height=171, Circumference=76, Age=30)
print(newObs)
```

$$
\begin{split}
\widehat{y} &= -66.579 + 0.4768\cdot171 + 0.7769\cdot76 - 0.2094\cdot30 \\
& = 67.72
\end{split}
$$

```{r, echo=TRUE}
newobs <- data.frame('Height'=171, 'Circumference'=76, 'Age'=30)
predict.lm(LinearModel.4, newdata=newobs, level=0.95)
```

## Other intervals in Prediction output

Confidence interval for $E(y|x_1, \ldots, x_p)$

```{r, echo=TRUE}
predict.lm(LinearModel.4, newdata=newobs, 
           level=0.95, interval=c("confidence"))
```

Prediction interval for $y|x_1, \ldots, x_p$

```{r, echo=TRUE}
predict.lm(LinearModel.4, newdata=newobs, 
           level=0.95, interval=c("prediction"))
```

## Compare real and predicted values

Let's predict the 387 extra observations and compare with actual observations!

```{r, echo=FALSE}
pred <- predict(LinearModel.4, newdata=bodydata[-c(1:20),], 
  interval="prediction", level=0.95)
head(pred,12)
```

## Predicted vs observed - plot

```{r, echo=FALSE, out.width=700, out.height=400}
newWeight <- bodydata[-c(1:20),1]
plot(newWeight,pred[,1], xlab="Observed", ylab="Predicted")
curve(1*x, add=TRUE, col=2, lwd=2)
```

## Checking the prediction intervals

How many observations fell within the prediction intervals?

```{r, echo=TRUE}
largerthanlower <- (newWeight > pred[,2])
smallerthanupper <- (newWeight < pred[,3])
inInterval <- which(largerthanlower & smallerthanupper)
print(length(inInterval))
```

Proportion of observations within prediction intervals:

```{r, echo=TRUE}
length(inInterval)/length(newWeight)
```

What should we expect?
