---
title: "GLM - Part I: Logistic Regression"
author: "Aliaksandr Hubin"
date: "Spring 2026"
output: 
   ioslides_presentation: default
   beamer_presentation: default
   slidy_presentation: default
editor_options: 
  markdown: 
    wrap: 72
---

## Book chapters

-   **An Introduction to Statistical Learning (ISLR)**: Chapter 4
    (Classification). Key sections:
    -   **4.2** Why Not Linear Regression?
    -   **4.3** Logistic Regression (The Logistic Model, Estimating
        coefficients, Making Predictions, Multiple Logistic Regression)
-   **The R Book (Crawley)**: Chapter 13 (Generalized Linear Models).
    Key sections:
    -   **Introduction to GLMs** (Error structures, Linear Predictors,
        and Link functions)
    -   **Proportion Data** (Binomial Errors for grouped data)
    -   **Binary Response Variables** (Modeling individual 0/1 outcomes)

## GLM's - What for?

-   A bigger class of linear models

-   Most single-response linear models are GLM's

-   Allows non-normally distributed response

    -   Binary response
    -   Proportion response
    -   Count response
    -   Multinomial response
    -   And more...

## Today's topics

-   Motivating examples: why linear regression fails for binary data
-   Logistic regression (binary or proportion response)
-   Estimation of parameters and Inference
-   Application to data examples
-   Binary classification revisited

## Recall linear regression

From our earlier lectures, we have observed data with a continuous
response $y_i$ and predictors $x_{i1}, \dots, x_{ip}$.

The multiple linear regression model is:
$$ \mu(\boldsymbol{x}_i) = E(y_i) = \beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip} $$

where the observations $y_i$ are typically assumed to be normally
distributed around the mean $\mu(\boldsymbol{x}_i)$ with some variance
$\sigma^2$: $$ y_i \sim N(\mu(\boldsymbol{x}_i), \sigma^2) $$

What happens if the response variable $y_i$ is **binary**, meaning it
can only take values 0 or 1?

## What if the response is binary?

Imagine we want to predict if a client will default on a loan (
$y \in \{0,1\}$) based on their account Balance ($x\in R$).

If we fit a simple linear regression model:
$$ E(y_i) = \beta_0 + \beta_1 x_{i1} $$

We want the expected value $E(y_i)$ to represent the **probability**
$P(y_i = 1)$, right? However, a straight line can predict probabilities
less than 0 or greater than 1, which is not possible!

## What if the response is binary? (Plot)

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=8}
# Require ISLR for the Default dataset
if(!require(ISLR)) install.packages("ISLR", repos="http://cran.us.r-project.org")
library(ISLR)
default_num <- ifelse(Default$default == "Yes", 1, 0)
par(mfrow=c(1,2))
plot(Default$balance, default_num, pch="|", col="orange", ylab="Probability of Default", xlab="Balance", main="Linear Regression")
abline(lm(default_num ~ balance, data=Default), col="blue", lwd=2)
plot(Default$balance, default_num, pch="|", col="orange", ylab="Probability of Default", xlab="Balance", main="Logistic Regression")
curve(predict(glm(default_num ~ balance, data=Default, family=binomial), data.frame(balance=x), type="response"), add=TRUE, col="blue", lwd=2)
```

## Modeling the Response

In Linear Regression, we assumed the continuous response $y_i$ followed
a **Normal Distribution** centered around its mean $\mu_i$:
$$ y_i \sim N(\mu_i, \sigma^2) $$

For a binary response, the data can only be $0$ or $1$. This perfectly
matches the **Bernoulli Distribution**!

The Bernoulli distribution describes a single trial that results in a
"success" ($y=1$) with probability $p_i$, or "failure" ($y=0$) with
probability $1-p_i$: $$ y_i \sim \text{Bernoulli}(p_i) $$

Our "mean" prediction is now simply the probability: $E(y_i) = p_i$.

## Normal vs Bernoulli Distribution

```{r, echo=FALSE, fig.height=4.5, fig.width=8}
par(mfrow=c(1,2), mar=c(4,4,3,1))

# Normal (Continuous)
x_norm <- seq(-3.5, 3.5, length=100)
y_norm <- dnorm(x_norm, mean=0, sd=1)
plot(x_norm, y_norm, type="l", lwd=3, col="blue", ylab="Density", xlab="y", main="Normal Distribution\n(Continuous y)", xaxt="n")
axis(1, at=0, labels=expression(mu))
polygon(c(x_norm, rev(x_norm)), c(rep(0, length(x_norm)), rev(y_norm)), col=rgb(0,0,1,0.2), border=NA)
abline(v=0, col="black", lty=2)

# Bernoulli (Binary)
y_bern <- c(0, 1)
prob_bern <- c(0.4, 0.6) # Example with p=0.6
plot(y_bern, prob_bern, type="h", lwd=15, col="red", ylab="Probability", xlab="y", main="Bernoulli Distribution\n(Binary y)", ylim=c(0, 1), xaxt="n", xlim=c(-0.5, 1.5))
axis(1, at=c(0,1), labels=c("0", "1"))
points(y_bern, prob_bern, pch=15, cex=3, col="red")
```

## Solution: Logistic regression

In logistic regression, we model the probability
$p_i = p(\boldsymbol{x}_i) = P(y_i = 1|\boldsymbol{x}_i)$ directly,
ensuring it stays between 0 and 1.

Instead of a straight line, we use the **logistic function** (or sigmoid
function):

$$ p(\boldsymbol{x}_i) = \frac{e^{ \beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip} }}{1 + e^{ \beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip} }} $$

This ensures that no matter what values the predictors $x$ take, the
predicted probability $p_i$ is always theoretically constrained:
$0 \leq p(\boldsymbol{x}_i) \leq 1$.

## The Logistic Function Shape

```{r, echo=FALSE, fig.height=4, fig.width=6}
x <- seq(-10, 10, by=0.1)
plot(x, exp(x)/(1+exp(x)), type="l", col="blue", lwd=2, ylab="p(x)", xlab="x", main="Standard Logistic Function")
lines(x, exp(0.5*x)/(1+exp(0.5*x)), col="red", lwd=2, lty=2)
lines(x, exp(2*x)/(1+exp(2*x)), col="green", lwd=2, lty=3)
legend("topleft", legend=c("Beta = 1", "Beta = 0.5", "Beta = 2"), col=c("blue", "red", "green"), lty=c(1,2,3), lwd=2)
```

Notice how it "squashes" the predictions to stay within 0 and 1.

## Log-odds (Logit) connection

How does logistic regression connect to linear regression?

If we rearrange the logistic formula, we find that the **log-odds** is
modeled linearly!

$$ \text{Odds} = \frac{p(\boldsymbol{x}_i)}{1 - p(\boldsymbol{x}_i)} = \frac{P(y_i=1|\boldsymbol{x}_i)}{P(y_i=0|\boldsymbol{x}_i)} $$

The logarithm of the odds is called the **logit** function:
$$ \log\left(\frac{p(\boldsymbol{x}_i)}{1 - p(\boldsymbol{x}_i)}\right) = \beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip} $$

So, a generalized representation uses a *link function* — here, the
logit link connects the mean (probability $p(\boldsymbol{x}_i)$) to the
linear predictor: $\beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip}$

## Interpretation of Logistic Coefficients

In a linear regression model, a coefficient $\beta_1$ means an increase
of 1 in $x_1$ leads to a $\beta_1$ increase in $E(y)$.

**What does** $\beta_1$ mean in Logistic regression?

Looking at the log-odds equation:
$$ \log(\text{Odds}) = \beta_0 + \beta_1 x_{1} $$

If we increase $x_1$ by one unit ($x_1 + 1$):

\- The **log-odds** increases by $\beta_1$.

\- The **odds** are multiplied by $e^{\beta_1}$.

$e^{\beta_1}$ is called the **Odds Ratio**.

## Parameter Estimation: Maximum Likelihood

In linear regression, we used the Least Squares method (which is
equivalent to Maximum Likelihood for normally distributed errors) to
find the best $\beta$ coefficients.

For Logistic Regression, there is no analytical "formula" (like the
least squares equations) for the coefficients.

Instead, the computer uses numerical algorithms to iteratively find the
$\beta$ values that maximize the **Likelihood function**.
$$ L(\boldsymbol{\beta}) \rightarrow \text{max} $$

Maximizing the likelihood means finding the parameter values that make
our observed binary data *most probable*.

## Likelihood Factorization

Since observations are independent, the total probability (Likelihood)
of seeing our data is the product of individual probabilities.

For a Binomial/Bernoulli outcome $y_i \in \{0, 1\}$, the probability is
$p_i$ if $y_i=1$, and $(1-p_i)$ if $y_i=0$.

We can write this elegantly as:
$$ L(\boldsymbol{\beta}) = \prod_{i=1}^n p_i^{y_i} (1 - p_i)^{1 - y_i} $$

We want to find $\boldsymbol\beta$ that maximizes this product (or its
sum of logs!).

## How the computer finds the Maximum

Instead of a set formula, the computer takes "steps" uphill on the
Log-Likelihood curve to find the peak!

```{r, echo=FALSE, fig.height=5, fig.width=6}
beta_vals <- seq(-3, 3, 0.1)
ll_vals <- -beta_vals^2 # Simple parabola mimicking log-likelihood
plot(beta_vals, ll_vals, type="l", lwd=2, ylab="Log-Likelihood", xlab=expression(beta), main="Newton-Raphson Intuition", ylim=c(-10, 2))
points(-1.5, -2.25, pch=16, col="blue", cex=1.5)
points(0, 0, pch=17, col="darkgreen", cex=2)
text(0, 1.2, "Maximum Likelihood Estimate (MLE)", col="darkgreen")
```

## The Wald Test

Just like linear regression, the maximum likelihood estimates
$\widehat{\beta}_j$ have standard errors, $SE(\widehat{\beta}_j)$.

For large sample sizes, $\widehat{\beta}_j$ approximately follows a
normal distribution.

We can test the hypothesis for individual coefficients
($H_0: \beta_j = 0$ vs $H_A: \beta_j \neq 0$) using the **Wald test**:
$$ z = \frac{\widehat{\beta}_j - 0}{SE(\widehat{\beta}_j)} $$

This statistic $z$ follows a Standard Normal distribution $N(0,1)$.
Software (like R) will output these $z$-values and corresponding
p-values.

## Standard Errors from MLE Theory

How does the computer know the Standard Error ($SE$) of the estimates?

It uses the **curvature** of the log-likelihood peak!

-   **Sharp peak:** We are very confident in the estimate. $SE$ is
    small.
-   **Flat peak:** There is a lot of uncertainty. $SE$ is large.

The mathematical term is the **Fisher Information** ($I$), which
measures the steepness (second derivative) of the peak.

$$ SE(\widehat{\beta}) = \sqrt{\frac{1}{I(\widehat{\beta})}} $$

So, sharp curvature = large Information = small Standard Error!

**The Motivation:** We want our proposed predictors to significantly
reduce the Lack of Fit. Therefore, we want the **Residual Deviance** to
be much smaller than the **Null Deviance**!

## Likelihood Ratio Test (Deviance)

To test if a subset of predictors as a whole is useful, we compare
nested models using the **Likelihood Ratio Test (LRT)**.

This is analogous to comparing the Sum of Squared Errors (SSE) in ANOVA
for linear models. Instead of SSE, we use **Deviance:**
$-2\log L(\boldsymbol{\beta})$.

-   The Null Model contains only the intercept $\beta_0$.
-   The Proposed Model contains the predictors.

The test compares the Deviance of the models, and the difference in
Deviances follows a $\chi^2$ (Chi-square) distribution based on the
number of additional predictors. In R, apply the `anova()` function with
`test="LR"` or `test="Chisq"` to see this.

## Likelihood Ratio Test (Intuition)

The Deviance compares the lack of fit of two models. If the Proposed
Model is **much better**, its deviance is much lower, making the
difference $\Delta \text{Deviance}$ large!

-   **Small difference:** The extra predictors don't do much.
-   **Large difference:** Pushes into the right tail (Rejection Region).
    The predictors are useful!

```{r, echo=FALSE, fig.height=3.5, fig.width=6}
x <- seq(0, 15, 0.1)
y <- dchisq(x, df=2)
plot(x, y, type="l", lwd=2, col="blue", ylab="Density", xlab=expression(Delta ~ Deviance ~ (chi^2)), main="LRT Statistic under the Null")
crit <- qchisq(0.95, df=2)
polygon(c(x[x>=crit], max(x), crit), c(y[x>=crit], 0, 0), col="red", border=NA)
abline(v=crit, col="black", lty=2, lwd=2)
text(8, 0.05, "Rejection Region\n(Significant predictors)", col="red", pos=4)
text(2, 0.15, "Do Not Reject\n(Small diff)", col="blue")
```

## Assumptions of Logistic Regression

Like linear models, GLMs have underlying assumptions we must check:

1.  **Independence:** Observations $y_i$ are independent of each other.
2.  **Linearity in the Logit:** The log-odds (logit) relates *linearly*
    to the continuous predictors.
3.  **No Multicollinearity:** Predictors shouldn't be too perfectly
    correlated.
4.  **No Extreme Outliers:** No single observation should unilaterally
    drive the model.
5.  **No Overdispersion** (for proportion data): The observed variance
    shouldn't be higher than what the Binomial theory assumes.

## Example 1: Bird species on islands

Let's test this in R! We look at the probability of a bird species being
present on an island.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(BIAS.data)
data(island)
head(island)
```

-   incidence = 1 if the species is present, 0 if absent
-   area = Area of the island ($km^2$)
-   isolation = Distance to the mainland ($km$)

Question: How does the probability of presence depend on the area?

## Fitting the Logistic Model in R

We use the `glm()` function, specifying `family=binomial(logit)` for
logistic regression. `mixlm` package is used over the standard base R
because it provides specific expanded summaries and visualizations in
the context of the course, or we can just use the base `glm`.

```{r, echo=TRUE, eval=FALSE}
library(mixlm)
GLM.1 <- glm(incidence ~ area, family=binomial(logit), data=island)
summary(GLM.1)
```

## The output

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(mixlm)
GLM.1 <- glm(incidence ~ area, family=binomial(logit), data=island)
summary(GLM.1)
```

## Interpreting the output

-   Intercept $\widehat{\beta}_0 = -2.16$
-   Area coefficient $\widehat{\beta}_{\text{area}} = 0.63$. The
    positive coefficient means larger area increases the odds of bird
    presence!
-   The **p-value** is $<0.001$, found using the Wald test (z value =
    $3.370$). The coefficient is significantly different from 0.
-   **Odds ratio**: $e^{0.63} \approx 1.872$. For every 1 $km^2$
    increase in area, the odds of the species being present on the
    island increase by about $87.2\%$.

## Graphical Representation of the Fit

```{r, echo=TRUE, fig.height=4, fig.width=6}
xv <- seq(0,9,0.01)
yv <- predict(GLM.1, list(area=xv), type="response")
plot(island$area, island$incidence, ylab="Probability of presence", xlab="Island Area")
lines(xv, yv, col="red", lwd=2)
```

## Goodness of fit (LRT)

We compare the model with the Area against the `Null`

```{r, echo=TRUE, eval=TRUE}
anova(GLM.1, test='LR')
```

## Goodness of fit (LRT) cont.

The Likelihood Ratio test from the previous output shows a significant
result: p-value\<$0.05$.

This indicates that adding the `area` predictor significantly improves
the fit of the model compared to a simple baseline model with just the
intercept.

## Expanding the model

What if we also include isolation?

```{r, echo=FALSE}
GLM.2 <- glm(incidence ~ area + isolation, family=binomial(logit), data=island)
summary(GLM.2)
```

Isolation has a negative coefficient: larger distance from mainland
decreases the probability of the species being present!

## Goodness of fit (LRT)

Comparing the models progressively:

```{r, echo=FALSE}
anova(GLM.2, test='LR')
```

Both variables give significant drops in deviance.

## Comparing Models: The AIC

The **Akaike's Information Criterion (AIC)** is commonly used to compare
models. It is defined as:

$$-2 \times \log L(\boldsymbol{\beta}) + 2 \times p$$

where $\log L(\boldsymbol{\beta})$ is the log-likelihood of the model
and $p$ is the number of estimated parameters. It penalizes complex
models. **Smaller is better**.

## Comparing Models: The AIC (Results)

```{r, echo=TRUE}
AIC(GLM.1) # Area only
AIC(GLM.2) # Area + Isolation
```

Adding isolation improves the model significantly, leading to a lower
AIC!

## Comparing Models: BIC

Along with AIC, we also use the **Bayesian Information Criterion
(BIC)**: $$ -2 \times \log L(\boldsymbol{\beta}) + \log(n) \times p $$
BIC penalizes complex models even more harshly than AIC because
$\log(n)$ is usually greater than $2$. Again, **smaller is better**.

```{r, echo=TRUE}
BIC(GLM.1) # Area only
BIC(GLM.2) # Area + Isolation
```

Both AIC and BIC agree: adding isolation improves the model!

## Checking the Bird Model (Diagnostics)

Let's verify our assumptions by generating all standard diagnostic plots
for `GLM.2`:

```{r, echo=TRUE, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
par(mfrow=c(2,2), mar=c(4,4,2,1))
plot(GLM.2)
```

## Binomial response

Sometimes, rather than tracking individual 0s and 1s, we have binomial
data.

$$ y_i \text{ successes out of } n_i \text{ trials}$$ For
$i = 1,\ldots, n$ where $y_i$ is the number of 'successes' among $n_i$
total experiments in group $i$.

We assume $y_i \sim \text{Binomial}(n_i, p_i)$.

## Binomial response (cont.)

Theory tells us: $$ E(y_i) = p_i $$
$$ Var(y_i) = \frac{p_i(1-p_i)}{n_i} $$

We can use logistic regression to model $p_i$ using predictors,
functioning precisely the same way!

## Example 2: The insect sex data

The sex counts in insects (proportion males) depending on population
density:

```{r, echo=FALSE}
load("sexratio.Rdata")
print(numbers)
```

## Binomial regression in R

In RStudio, the response for Binomial grouped data must be an array of
`successes` and `failures`.

In the model specification, use `cbind(succesess, failures)` as the
response variable.

```{r, echo=TRUE}
GLM.4 <- glm(cbind(males, females) ~ log(density), 
             family=binomial(logit), data=numbers)
```

## Insect sex-ratio model summary

```{r, echo=FALSE}
summary(GLM.4)
```

## Insect sex-ratio plot

```{r, echo=FALSE, fig.height=4, fig.width=6}
attach(numbers)
p <- males/(males+females)
xv<-seq(0,6,0.1)
plot(log(density), p, ylab="Proportion male")
lines(xv, predict(GLM.4, list(density=exp(xv)), type="response"), col="red", lwd=2)
detach(numbers)
```

## Binary classification revisited

Remember from Multivariate Analysis: Assume we have two groups
(classes), $c_1$ and $c_2$. By Bayes rule, the posterior probability of
class 2 given observed variables is:

$$ P(c_2 | x_1,\dots, x_k) \propto f(x_1,\dots, x_k | c_2) P(c_2) $$

Logistic regression gives an alternative, **direct** method to estimate
this posterior probability!

Define $y=1$ for $c_2$ and $y=0$ for $c_1$. Then logistic regression
directly estimates:
$$ P(c_2 | x_1,\dots, x_k) = P(y=1 | x_1,\dots, x_k) = \frac{e^{ \beta_0 + \beta_1 x_1 + \dots}}{1 + e^{ \beta_0 + \beta_1 x_1 + \dots}} $$

We classify to class 2 if the estimated $P(y=1 | x_1,\dots, x_k) > 0.5$.

## Example 3: Lizards

Can we discriminate the Sex of Lizards based on Mass? Let $y=1$ if
female, $y=0$ if male.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data(Lizard)
y <- ifelse(Lizard$Sex=="m", 0, 1)
model <- glm(y ~ Mass, family=binomial, data=Lizard)
summary(model)$coef
```

For every 1 unit increase in Mass, the log-odds of being female changes
by $-0.89$. Thus heavier lizards are more likely to be male!

## Lizards: Fitted line plot

```{r,echo=FALSE, fig.height=4, fig.width=6}
plot(Lizard$Mass, y, ylab="Probability of Female", xlab="Mass", pch=16)
x <- seq(2,16, 0.05)
ypred <- predict(model, list(Mass=x), type="response")
lines(x, ypred, col="red", lwd=2)
abline(h=0.5, lty=3)
```

## Classify a new lizard

Assume we have measured a new lizard with mass 5g. What is the most
probable gender for this lizard given our estimated model?

```{r,echo=TRUE}
new_lizard_data <- data.frame(Mass = 5)
ypred <- predict(model, new_lizard_data, type="response")
ypred
```

This returns the fitted estimated probability of being female ($y=1$).

Since $0.957 > 0.5$, we classify the lizard to the female class. Hence,
a lizard of Mass 5g is most probably a female.

## Lizards: Variable Selection

Let's include all predictors (`Mass`, `SVL` - Snout-vent length, `HLS` -
Hind limb span) and perform stepwise selection!

```{r, echo=TRUE, message=FALSE, warning=FALSE}
full_model <- glm(y ~ Mass + SVL + HLS, family=binomial, data=Lizard)
selected_model <- step(full_model, direction="both", trace=0,k = 2)
summary(selected_model)$coef
```

The stepwise method selected only `HLS` as the best predictor (lowest
AIC).

## Lizards: Out of sample evaluation

Let's measure how well our model really works by splitting data into a
**training set** and a **test set**.

```{r, echo=TRUE,warning=FALSE}
set.seed(42)
train_idx <- sample(1:nrow(Lizard), size = 0.6 * nrow(Lizard))
train_data <- Lizard[train_idx, ]
test_data <- Lizard[-train_idx, ]

# Fit on training data
train_model <- glm(ifelse(Sex=="m", 0, 1) ~ HLS, 
                   family=binomial, data=train_data)
```

## Lizards: Training sample accuracy

First, let's evaluate the model on the **training data**:

```{r, echo=TRUE}
train_probs <- predict(train_model, train_data, type="response")
train_preds <- ifelse(train_probs > 0.5, 1, 0)
train_actuals <- ifelse(train_data$Sex=="m", 0, 1)

# Training Confusion Matrix
table(Predicted = train_preds, Actual = train_actuals)
```

Training Accuracy: `r mean(train_preds == train_actuals)`

## Lizards: Testing sample accuracy

```{r, echo=TRUE}
test_probs <- predict(train_model, test_data, type="response")
test_preds <- ifelse(test_probs > 0.5, 1, 0)
test_actuals <- ifelse(test_data$Sex=="m", 0, 1)

# Test Confusion Matrix
table(Predicted = test_preds, Actual = test_actuals)
```

Test Accuracy: `r mean(test_preds == test_actuals)`
