---
title: "ANOVA 2"
author: "Aliaksandr Hubin"
date: "Mar 02 2025"
output: 
   ioslides_presentation: default
   beamer_presentation: default
   slidy_presentation: default
   output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

## Topics for this lecture

-   ANOVA with several factors
-   The factorial applet
-   Effect plots
-   Estimation and hypothesis testing
-   Tukey test and confidence intervals
-   ANCOVA - <Example:Regression> with indicator variables

## ANOVA with several factors

-   Each with two or more levels
-   Assume two factors: A with $i=1,2,...a$ levels, and B with
    $j=1,2,...,b$ levels
-   If all levels of A are tested with all levels of B we have a
    complete crossed design.
-   Assume there are $k=1,2,...,n$ replicates of all $ab$ level
    combinations.

## Illustration of two-factor case

<https://solve.shinyapps.io/Factorialdemo/>

## The Model

Model:
$$y_{ijk} = \mu + \alpha_i + \beta_j + \alpha\beta_{ij} + \epsilon_{ijk}$$

where $\epsilon_{ijk} \sim N(0,\sigma^2)$.

Parametrization (restrictions)

-   Sum-to-zero: $\sum_i {\alpha}_i =0$, $\sum_j {\beta}_j=0$,
    $\sum_i \alpha\beta_{ij}=0$ and $\sum_j \alpha\beta_{ij}=0$.

-   Reference levels: $\alpha_1 = 0$, $\beta_1 = 0$,
    $\alpha\beta_{1,j}=0$ for all $j$ and $\alpha\beta_{i,1}=0$ for all
    $i$.

## Example: Animal diets - The "growth data"

```{r,echo=FALSE, message=FALSE}
load("growth.Rdata")
attach(weights)
barplot(tapply(gain, list(diet, supplement),mean),beside=TRUE, ylim=c(0,35),
        col=c("orange","yellow","cornsilk"))
legend(3.7, 34.8, legend=c("Barley","Oats","Wheat"),fill=c("orange","yellow","cornsilk"))
```

## R analysis

For reference level analysis, see "the R-book" on page 517.

Let's again use the sum-to-zero parametrization here:

```{r, echo=TRUE, message=FALSE}
mod2 <- lm(gain ~ diet + supplement + diet:supplement, 
           data=weights, contrasts = "contr.sum")
```

Equivalent model calls in R:

-   `gain ~ diet + supplement + diet:supplement`
-   `gain ~ diet*supplement`
-   `gain ~ (diet + supplement)^2`

## Diet means and sd's

```{r, echo=TRUE, message=FALSE}
by(weights$gain, weights$diet, 
   FUN=function(x){c(mean=mean(x),sd=sd(x))})
```

## Supplement means and sd's

```{r, echo=TRUE, message=FALSE}
by(weights$gain, weights$supplement, 
   FUN=function(x){c(mean=mean(x),sd=sd(x))})
```

## Main effects plot of `Diet`

"Main effect plots" of group means with 95% CI's

```{r,echo=TRUE, message=FALSE}
library(effects)
plot(Effect(focal.predictors = c("diet"), mod = mod2))
```

## Main effects plot of `Supplement`

```{r,echo=TRUE, message=FALSE}
plot(Effect(c("supplement"), mod2))
```

## Interaction plots

Means of diet:supplement combinations.

```{r,echo=TRUE, message=FALSE}
library(effects)
plot(allEffects(mod2,confidence.level=.99),layout=c(3,1), 
     rotx=90,alternating=FALSE)
```

## Model parameter estimation

For balanced data the least squares estimators for the model parameters
are (sum-to-zero):

$\widehat{\mu} = \bar{y}_{...}$

$\widehat{\alpha}_i = \bar{y}_{i..}-\bar{y}_{...}$

$\widehat{\beta}_j = \bar{y}_{.j.}-\bar{y}_{...}$

$\widehat{\alpha\beta}_{ij} = \bar{y}_{ij.} - \bar{y}_{i..} - \bar{y}_{.j.} + \bar{y}_{...}$

And as usual, the noise variance estimate is:

$\widehat{\sigma}^2 = MSE$

## Summary with estimates of effects:

```{r, echo=FALSE}
summary(mod2)$coefficients[,c(1:2)]
```

## We want to test hypotheses

Interaction effects?
$$H_0:\alpha\beta_{11} = \alpha\beta_{12} = \cdots = \alpha\beta_{ab} = 0$$

Main effect of A? $$H_0:\alpha_1 = \alpha_2=\cdots=\alpha_a = 0$$

Main effect of B? $$H_0:\beta_1 = \beta_2 = \cdots = \beta_b = 0$$

## Splitting of sums of squares

The Total Sums of Squares measures total variability:
$$SST = \sum_i \sum_j \sum_k (y_{ijk} - \bar{y}_{...})^2$$

Can be split into: $$SST = SSA + SSB + SSAB + SSE$$

Degrees of freedom follow correspondingly as
$$(N-1) = [(a-1)] + [(b-1)] + [(a-1)(b-1)] + [ab(n-1)]$$

## F-tests

Interaction: Reject $H_0$ if
$$F = \frac{MSAB}{MSE} = \frac{SSAB/(a-1)(b-1)}{SSE/ab(n-1)}$$ is larger
than $F_{\alpha,(a-1)(b-1),ab(n-1)}$ or p-value \< $\alpha$.

Main effect A: Reject $H_0$ if
$$F = \frac{MSA}{MSE} > F_{\alpha,(a-1),ab(n-1)}$$

Main effect B: Reject $H_0$ if
$$F = \frac{MSB}{MSE} > F_{\alpha,(b-1),ab(n-1)}$$

## The ANOVA-table for the growth data

```{r, echo=TRUE, message=FALSE}
anova(mod2)
```

Conclusions?

## A reduced model

```{r, echo=TRUE}
mod3 <- lm(gain ~ diet + supplement, data=weights, contrasts = "contr.sum")
anova(mod3)
```

Note: The complete crossed design gives "orthogonal effects". The SS's
are independent, and, apart from SSE, they do not change when effects
are removed/added to the model.

## Tukey tests supplement - Confidence intervals

```{r, message=FALSE, echo=FALSE}
library(mixlm)
ptests <- simple.glht(mod3, "supplement", corr = c("Tukey"),level = 0.95)
print(ptests$confint)
```

## Compact letter display

A compact letter display (cld) of groups of effects

```{r, echo=TRUE}
cld(ptests) 
```

## Analysis of covariance - ANCOVA

The dataset `birth` in the BIAS.data package holds records of 189 birth
weights from Massachusetts, USA, and some additional variables. Three of
the variables are,

| Var   | Description                                |
|-------|--------------------------------------------|
| `LWT` | Weight of mother.                          |
| `SMK` | `YES` if mother smokes and `NO` otherwise. |
| `BWT` | Birth weight (RESPONSE).                   |

The main question is: Does mother's smoking affect birth weight and does
it depend on the mother weight?

We have both a categorical and a continuous predictor for the response
variable.

## The data

```{r, echo=FALSE}
library(BIAS.data)
data(birth)
head(birth[,-c(1:2)], 7)
```

## Scatterplot with separate linear models for smokers and non-smokers

```{r, echo=FALSE}
library(car)
scatterplot(BWT~LWT | SMK, regLine=lm, smooth=FALSE,
            boxplots=FALSE, ellipse=FALSE, by.groups=TRUE,data=birth)
```

## ANCOVA as regression with indicator variable

Let $y = BWT$ and $x_1 = LWT$ and define $$ 
x_2 = \left\{ 
\begin{array}[cc]
\ 0 & \mbox{if smoker}\\
1 & \mbox{if non-smoker}
\end{array} \right.
$$

Assume the model $$ 
y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \epsilon_i
$$ where $\epsilon_i \sim N(0, \sigma^2)$.

## Fitting the model in R

```{r, echo=TRUE, message=FALSE}
birth4 <- lm(BWT ~ LWT + SMK, data=birth, contrasts = "contr.treatment")
birth4
```

## Explain coefficients

## The estimated model(s)

```{r, echo=FALSE}
plot(x = birth$LWT, y = birth$BWT, xlab="LWT", ylab="BWT", col=ifelse(birth$SMK=="YES", 2,4))
coefs <- birth4$coefficients
abline(coefs[1], coefs[2], col = "red", lwd=2)
abline((coefs[1]+coefs[3]), coefs[2], col = "blue", lwd=2)
```

## The model with interaction term

Assume that we expand the model with an interaction term between LWT and
SMK $$ 
y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 x_{1i}\cdot x_{2i} + \epsilon_i
$$ with $\epsilon_i \sim N(0, \sigma^2)$.

Lets see how the model is for different values of $x_{2i}$

## Explain coefficients

## The estimated model(s)

```{r, echo=FALSE}
plot(x = birth$LWT, y = birth$BWT, xlab="LWT", ylab="BWT")
birth5 <- lm(BWT ~ LWT*SMK, data=birth, contrasts="contr.treatment")
coefs <- birth5$coefficients
abline(coefs[1], coefs[2], col = "red", lwd=2)
abline((coefs[1]+coefs[3]), (coefs[2]+ coefs[4]), col = "blue", lwd=2)
```

## Model summaries, tests, diagnostics

You should be able to check:

-   Are model assumptions met?
-   Are there significant effects of LWT and/or SMK?
-   How is the model fit in terms of $R^2$? Critizise the model!
