---
title: "ANOVA 1"
author: "Aliaksandr Hubin"
date: "Feb 24 2025"
output: 
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
  # output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

## Topics for this lecture

-   Introductory example
-   The "ANOVA" shiny-applet
-   The F-test
-   Estimation and parametrization
-   Post hoc contrasts and pairwise tests
-   Errors and power

## Analysis of variance (ANOVA)

**Regression models**

-   Response is numerical
-   Predictors are numerical, some may be factor (ANCOVA)

**Analysis Of Variance (ANOVA)**

-   Response is numerical
-   Predictors are all factors (categorical), possibly with many levels

**Fitting the model to data (maximum likelihood) is the same**

**Explanation**

-   Are there any effect of the predictors on the response?

## Example: The "Yields" data

Yield from a crop experiment on $k=3$ different soil types. There are
$n=10$ random replicates per soil type. A total of $N=k\cdot n = 30$
observations altogether.

A box-plot. Are the expected yields different?

```{r, echo=FALSE, message=FALSE}
library(BIAS.data)
library(mixlm)
data(yields)
attach(yields)
plot(yield ~ soil, data=yields, col="green")
```

## The model

As in regression analysis we assume a linear model with both an
explainable part and an unexplainable part (noise)

Let $y_{ij}$ be the observed yield for observation number
$j \in \{1,\ldots, n_i\}$ soil type $i \in \{1,\ldots, k\}$.

The model (One-way ANOVA): $$y_{ij} = \mu_{i} + \epsilon_{ij}$$

where $\epsilon_{ij} \sim N(0, \sigma^2)$ and $\mu_i$ is the expected
yield for soil type $i$.

## Hypotheses testing in ANOVA

If there is a soil effect on yield, the expected yields will differ in
one way or the other.

Hypotheses:

$$H_0: \mu_1 = \mu_2 =\mu_3 $$ vs $$H_1: \mu_i \neq \mu_{i'}$$ for at
least two different soil types $i$ and $i'$.

Recall F-test in regression? Does it look familiar?

## The ANOVA applet

The ANOVA "idea": Studying means by splitting variability

[ANOVA shiny app](https://solve.shinyapps.io/ANOVA/)

## Splitting up total variability

-   Total variation in $y$: SST=
    $\sum_{i=1}^{k}\sum_{j=1}^{n}(y_{ij}-\bar{y}_{..})^2$

For the yield data:

```{r, echo=FALSE}
sum((yield-mean(yield))^2)
```

-   Variation between levels: SSA =
    $\sum_{i=1}^{k}\sum_{j=1}^{n}(\bar{y}_{i.}-\bar{y}_{..})^2 =
    n\sum_{i=1}^{k}(\bar{y}_{i.}-\bar{y}_{..})^2$

For the yield data:

```{r, echo=FALSE}
soilmeans <- by(yield, soil, mean)
10*sum((soilmeans - mean(yield))^2)
```

-   Variation within levels: SSE = SST - SSA =
    $\sum_{i=1}^{k}\sum_{j=1}^{n}(y_{ij}-\bar{y}_{i.})^2$ = 414.7 - 99.2
    = 315.5

## Between and within group variances:

Dividing Sums of Squares by their respective degrees of freedom gives:

-   The between soil type (group) variance:
    $MSA = SSA/(k-1) = 99.2/2 = 49.6$
-   The within soil type (group) variance:
    $MSE = SSE/(N-k) = 315.5/27 = 11.685$

where $N = n_1 + n_2 + \cdots n_k$ is the total number of observations.

Why does it make sense to compare these?

## Testing the effect of soil

If there is a soil effect on yield, the expected yields will differ in
one way or the other.

Hypotheses:

$$H_0: \mu_1 = \mu_2 =\mu_3 $$ vs $$H_1: \mu_i \neq \mu_{i'}$$ for at
least two different soil types $i$ and $i'$.

Reject $H_0$ if $MSA >> MSE$

## Test observator, F

If $H_0$ is true, then

$$F = \frac{MSA}{MSE} \sim F(\nu_1, \nu_2)$$

```{r, echo=FALSE}
curve(df(x,2, 27), xlim=c(0,7), lwd=2, col=2, ylab="Fisher(2,27)", xlab="F")
abline(h=0, v=c(0, 4.24))
text(5.5, 0.2, labels="F = 49.6/11.7 = 4.24")
```

## Rejection region

Reject $H_0$ at test level $\alpha$ if
$F_{obs}> F_{\alpha, \nu_1, \nu_2}$, where $\nu_1$ are the degrees of
freedom for MSA and $\nu_2$ are the degrees of freedom for MSE.

For example, assume a test level of 5%

From a Fisher table we find $F_{0.05, 2, 27}$ = 3.35

Alternatively, reject if p-value from R is smaller than $\alpha$.

Conclusion?

## Estimation of group effects and error variance

The inutitive estimators for $\mu_i$ are
$\widehat{\mu}_i = \bar{y}_{i.}$

These are also the least squares estimators (for balanced data), which
are unbiased and have minimum variance among all unbiased estimators.

As in regression analysis we have $\widehat{\sigma}^2 = MSE$.

## Splitting up $\mu_i$

The means may be split up further. Assume

$$\mu_i = \mu + \alpha_i$$

But, all parameters can not be estimated without some extra restriction
(choice of parametrization).

-   Parametrization 1: Assume $\sum_{i=1}^k \alpha_i = 0$

A "sum to zero" parametrization ("contr.sum" in R")

-   Parametrization 2: Assume $\alpha_1 = 0$

A "reference level" or "control level" parametrization
("contr.treatment" in R")

## Draw all parametrizations

## Choice of arametrization in RStudio

Parametrization 1 (Using the lm-function in the mixlm-package)

```{r, echo=TRUE, eval=FALSE}
mod1 <- lm(yield~soil, data=yields, contrasts = "contr.sum")
```

Parametrization 2 (Using the lm-function in the mixlm-package

```{r, echo=TRUE, eval=FALSE}
mod1 <- lm(yield~soil, data=yields, contrasts = "contr.treatment")
```

## Example: Parametrization 1

```{r, echo=FALSE}
mod1 <- lm(yield~soil, data=yields, contrasts = "contr.sum")
round(summary(mod1)$coef,4)
```

Compare with group and total means:

```{r}
means <- c(soilmeans,mean(soilmeans))
names(means) <- c("clay", "loam", "sand", "total-mean")
print(means)
```

## Example: Parametrization 2

```{r, echo=FALSE}
mod1 <- lm(yield~soil, data=yields, contrasts = "contr.treatment")
round(summary(mod1)$coef,4)
```

Compare with group and total means:

```{r, echo=FALSE}
means <- c(soilmeans,mean(soilmeans))
names(means) <- c("clay", "loam", "sand", "total-mean")
print(means)
```

## ANOVA table of Sum of squares

This table does not depend on parametrization

```{r, echo=FALSE}
anova(mod1)
```

## Residual analysis

We should also for ANOVA-models check the assumptions by residual
analysis

```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(mod1, which=c(1,2))
```

## All pairwise comparisons

-   We want to test if there are differences **between the levels**
    -   Is the yield for `clay` significantly different from the yield
        for `sand`?
    -   Is the yield for `clay` significantly different from the yield
        for `loam`?
    -   Is the yield for `loam` significantly different from the yield
        for `sand`?
    -   ...with $k$ factor levels there are $k(k-1)/2$ pairs...
-   **Correcting for multiple testing**
    -   With a 5% rejection level, we will on average in 5% of the tests
        claim falsely significance (reject $H_0$)
    -   When making **many** tests, we should be (slightly) stricter
-   **We perform pairwise tests with Tukey correction!**

## Pair-wise contrasts

If a factor is significant and has more than two levels: Where are the
differences?

May perform a "post hoc" test with contrasts.

A contrasts is a linear function of the factor effects.

$$\theta = c_1 \mu_1 + c_2 \mu_2 + c_3 \mu_3$$

where $\sum_{i=1}^k c_i= 0$

For pair-wise contrasts, only two $c_i$'s differ from 0.

## Example 1 

Comparing soil type "loam" (level 1) with "clay" (level 2):

$\theta_1 = 1\mu_1 + (-1)\mu_2 + 0 \mu_3 = \mu_1 -\mu_2$

$= 1(\mu + \alpha_1) - 1(\mu + \alpha_2) = \alpha_1 -\alpha_2$

Estimate:
$\widehat{\theta}_1 = \widehat{\mu}_1 -\widehat{\mu}_2 =  11.5 - 14.3 = \widehat{\alpha}_1 -\widehat{\alpha}_2 = -0.4 - 2.4 = -2.8$

```{r, echo=TRUE}
library(gmodels)
options(contrasts=c("contr.sum","contr.poly")) 
mod1 <- lm(yield~soil, data=yields)
gmodels::fit.contrast(model=mod1, varname='soil', df=TRUE, coeff=c(1, -1, 0), conf.int=0.95)
```

## Example 2: "loam" vs "sand"

Define required pairwise contrast:

$\theta_2 = 0\mu_1 + 1\mu_2 - 1\mu_3$
$= 1(\mu + \alpha_2) - 1(\mu + \alpha_3) = \alpha_2 - \alpha_3$

Estimation

$\widehat{\alpha}_3 = -(-0.4 + 2.4) = -2$
$\widehat{\theta}_2 = \widehat{\alpha}_2 - \widehat{\alpha}_3 = 2.4 - (-2) = 4.4$

```{r, echo=FALSE}
gmodels::fit.contrast(model=mod1, varname='soil', df=TRUE, coeff=c(0, 1, -1), conf.int=0.95)
```

## Multiple testing - Tukey test

If there are many levels of a factor, a Tukey test is preferable.

Controls overall error rate (family-wise error) of all pairwise tests.

Use 'simple.glht'-function in the mixlm-package.

```{r, echo=TRUE, eval=FALSE}
ptests <- simple.glht(mod1, "soil", corr = c("Tukey"),level = 0.95)
print(ptests)
```

## R Commander output - Tukey

```{r, echo=FALSE}
ptests <- simple.glht(mod1, "soil", corr = c("Tukey"),level = 0.95)
print(ptests)
```

## A compact letter display of effects

```{r, echo=FALSE}
cld(simple.glht(mod1,'soil', level=0.95))
```

## Pairwise confidence intervals plot

```{r, echo=FALSE}
par(mar=c(5,6,4,1))
plot(ptests)
```

## Type 1 and 2 errors

We should keep in mind the errors which we may make

Type 1: $\alpha$ = P(Rejecting $H_0$ \| $H_0$ is TRUE)

Type 2: $\beta$ = P(Retaining $H_0$ \| $H_0$ is FALSE)

Choosing the test level $\alpha$ always has an impact on $\beta$.

## Test power

The test power is the ability to find existing effects:

Power: $(1-\beta)$ = P(Rejecting $H_0$ \| $H_0$ is FALSE)

Which factors influence the power for a given test?

Back to: [ANOVA shiny app](https://solve.shinyapps.io/ANOVA/)
