---
title: "Exercises STAT340 - ANOVA - part I"
params:
  vis.forslag: false
output:
  html_document:
    toc: true
    theme: yeti
    code_folding: hide
    number_sections: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style>
.main-container, .main-content { max-width: 100%; }
.info { background-color:aliceblue; }
.task { background-color:lightpink; }
</style>
```

```{r forslag, include = FALSE}
vis.forslag <- params$vis.forslag
```

```{r forklar forslag, echo = FALSE, results = "asis", include = vis.forslag}
cat('
<mark>
**LØSNINGSFORSLAG!**

Disse vises nedenfor på ulike måter:
<mark>Gulmerking</mark> av tekstforslag til løsning, eller kommentarer om alternative løsninger.
* Noe <button>Code</button> som var skjult i øvingsdokumentet.
<details>
<summary>
<mark>
Noen løsningsforslag er skjult inni hint.
</mark>
</summary>
Da må du utvide hintene for å se løsningsforslagene.  

</details>
')
```

<br> <br> <br> <br>

::: info
**Colloquium exercises:** You are encouraged to work together in groups
on the exercises. Sharing problems and solutions is often a much quicker
way to learn, especially for problems of a more practical nature.
:::

<br>

```{r, echo = F, message=FALSE}
library(BIAS.data)
library(mixlm)
data(city)
```

# Ex-1: Chlorine levels in cities

Independent measurements of chlorine (ppm parts per million) were taken
from 3 large cities:

```{r, echo=FALSE}
city
```

-   Load the city dataset from the BIASdata R-package.

Before analyzing the data it is best to "stack" the data into two
columns, a response column "y" and a city factor column "city". This can
be done using the function stack() in RStudio. You can name your data
set as "citydata", the response variable as "y" and the factor as
"city". Try to do this by the function colnames().

<details>

<summary><mark> Solution </mark></summary>

Rstudio:

```{r, echo=TRUE}
citydata <- stack(city)
colnames(citydata)<-c("y", "city")
```

The top of your data should look like this:

```{r, echo=FALSE}
head(citydata, 10)
```

</details>

-   Assume an ANOVA-model with chlorine as response and city as a
    factor. What assumptions do you make?

<details>

<summary><mark> Solution </mark></summary>

The model assumptions are: $$y_{ij} = \mu_i + \epsilon_{ij}$$ where the
error terms are assumed independent and identically distributed
$\epsilon_{ij} \sim N(0, \sigma^2)$. Further $\mu_i$ is the expected
chlorine level in city $i$.

</details>

-   Make a box-plot of the data using the Boxplot-function in the
    'car'-package Describe the variabilities between cities and within
    cities.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
library(car)
Boxplot(y~city, data=citydata, id.method="y")
```

City 1 seems to have both the lowest average chlorine levels and the
smallest variability in the measurements. Cities 2 and 3 are quite
similar with both higher means and variability.

</details>

-   Fit the model (using sum to zero parametrization) and perform a
    residual analysis. Comment on the model fit.

<details>

<summary><mark> Solution </mark></summary>

The fitted model:

```{r, echo=TRUE}
options(contrasts=c("contr.sum", "contr.poly"))
citymod1 <- lm(y ~ city, data=citydata)
summary(citymod1, cor=FALSE)
anova(citymod1)
par(mfrow=c(2,2))
plot(citymod1, which=c(1,2,4,5))
```

The residual analysis reveals two problems: 1) Non-constant variability,
which was also observed in the boxplot, and 2) A skeewed and non-normal
distribution for the residuals with a heavy tail to the right. The $R^2$
is quite small, only 16.4%, so there is a lot of unexplained variation
in the response variable. The hypothesis test of effect of the group
variable `city` appears to be highly significant, but we should be
careful conluding about the effect of city until we have tried to
improve the model fit.

</details>

-   Make a log-transformation of $y$ to create the variable `logy`

```{r, echo=TRUE}
citydata$logy <- log(citydata$y)
```

-   Fit a new model with `logy` as response and check the model fit
    again.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=FALSE}
citymod2 <- lm(logy ~ city, data=citydata)
summary(citymod2, cor=FALSE)
anova(citymod2)
par(mfrow=c(2,2))
plot(citymod2, which=c(1,2,4,5))
```

Both problems from the previous model appear to be corrected. The $R^2$
has also improved a bit. We continue with this model.

</details>

-   Estimate all model parameters from the latter model. Explain what
    the parameters measure given the paramtrization you have used.

<details>

<summary><mark> Solution </mark></summary>

The ANOVA analysis using sum-to-zero parametrization gave the following
results.

```{r, echo=TRUE}
coefs <- citymod2$coefficients
```

We observe that $R^2 = 0.25$ indicating that there is still a lot of
unexplained variability in the response. We get the following estimates
for the city means of log(chlorine)-levels using this parametrization.
The $\mu$ can be interpreted as a general expected log(chlorine)-level
for all three cities, which is estimated to be
$\hat{\mu}$=`r round(coefs[1],2)` corresponding to
$exp(\hat{\mu})$=`r round(exp(coefs[1]),2)` on the original scale of
chlorine level. Further, city-specific means are estimated as (on th e
log-scale):

$\hat{\mu}_1 = \hat{\mu} + \hat{\alpha}_1 =$
`r round(coefs[1] + coefs[2],2)`

$\hat{\mu}_2 = \hat{\mu} + \hat{\alpha}_2 =$
`r round(coefs[1] + coefs[3],2)`

$\hat{\mu}_3 = \hat{\mu} + \hat{\alpha}_3 = \hat{\mu} -(\hat{\alpha}_1 + \hat{\alpha}_2)=$
`r round(coefs[1] -(coefs[2]+coefs[3]),2)`

The estimate of the error variance is
$\hat{\sigma}^2 = MSE= SSE/(n-3) = 60.724/57 = 1.065$.

The $\sigma^2$ is the variance for log-chlorine levels measured in the
same city.

</details>

-   State the hypotheses for testing whether the expected chlorine
    levels differ between the cities, choose a test level $\alpha$ and
    perform the test. What is the conclusion?

<details>

<summary><mark> Solution </mark></summary>

The hypotheses are: $$H_0: \mu_1 = \mu_2 = \mu_3$$ versus
$$H_1:\mu_i \neq\mu_{i'}$$ for at least two different cities $i$ and
$i'$. Note:This is by the chosen parametrization equivalent to testing
$$H_0: \alpha_1 = \alpha_2 = \alpha_3 = 0$$ versus $$H_1:\alpha_i\neq0$$
for at least one $i \in \{1,2,3\}$. For test-level $\alpha=0.05$ we
reject the null-hypothesis if $F>F_{\alpha, 2, 57}$ or if the p-value is
smaller than 0.05. Here we observe a very small p-value and reject the
null-hypothesis. We claim that the expected log-chlorine levels differ
between at least two of the cities, and from the observed means we know
that cities 1 and 3 are significantly different since they have the
largest observed difference in means. There may also be a significant
difference between cities 1 and 2, and 2 and 3, but this should be
checked by a pair-wise contrast.

</details>

-   Test the pairwise differences between the cities using the Tukey
    test. Which cities are significantly different from each other?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
print(simple.glht(citymod2,'city', level=0.95))
cld(simple.glht(citymod2,'city', level=0.95))
```

From the Tukey-corrected p-values (Last column of the Linear hypotheses
table) we observe that City 1 and 2 are sognificantly different (at 5%
test level). The same goes for City 1 and 3. City 2 and 3 are not
significantly different. This can also quickly be seen from the compact
letter display at the bottom. From the mean values we see that City 1
has a significantly lower chlorine level than the two other cities.

</details>

# Ex-2: Data from the NSR education test

The Norwegian Centre for Science Recruitment (NSR) has an online
"education test" where youths may answer a questionnaire to check their
so-called cognitive types, their science interest, their preferred
learning methods and their interest to various science subjects. The
test suggests different areas within the STEM (Science, Technology,
Engineering and Mathematics) within which the youth may find suitable
work.

We have an excerpt of these data which can be downloaded from Fronter as
the NSR.Rdata file. The data.frame `NSRdata` contains two variables,
Science and Age (first 5 observations):

```{r, echo=FALSE}
data(NSRdata)
head(NSRdata,5)
```

"Science" is an average liking score (scale 1-6) to various
STEM-subjects, and "Age" is a factor indicating different age-groups:

-   1: 1-12 yrs

-   13: 13-15 yrs

-   16: 16-19 yrs

-   19: 19-29 yrs

-   30: 30 + yrs

-   Perform an analysis of the NSR data to check whether Age influences
    the liking to STEM subjects. State the model, fit the model, check
    model assumptions, test hypotheses, and give model critique. Write a
    short summary of the results.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
NSRmod <- lm(Science ~ Age, data = NSRdata)
anova(NSRmod)
pt <- simple.glht(NSRmod, "Age", corr = c("Tukey"),level = 0.95)
print(pt)
```

The anova table indicates a clear significance of Age-group with regard
to the interest level to Science subjects. Hence, at least two
age-groups are have significantly different expected means. We can use
the Tukey test to identify pairwise differences.

The Tukey output provides differences in mean (column "center") between
all pairs of groups, and at the top of output the minimum difference
yielding a significant difference in means is given to be
`r unique(round(with(pt$res, (Upper - Lower)/2), 4))`. That is, all
age-groups with a difference in averages of more than
`r unique(round(with(pt$res, (Upper - Lower)/2), 4))` are significantly
different, according to Tukey. From the p-values we observe that
`r sum(pt$res[,6] < 0.05)` out of `r dim(pt$res)[1]` pairs are
significantly different, and the largest difference is found between age
groups 30+ and 19-29.

```{r, echo=TRUE}
cld(pt)
```

The compact letter display gives a grouping of the similar levels, and
there are three groups of levels that are internaly non-significantly
different. 30+ is different from all other levels, whereas 1-12 and
16-19 are similar and 16-19, 13-15 and 19-29 are also similar.

</details>
