---
title: "Exercises STAT340 - ANOVA - part II"
params:
  vis.forslag: false
output:
  html_document:
    toc: true
    theme: yeti
    code_folding: hide
    number_sections: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style>
.main-container, .main-content { max-width: 100%; }
.info { background-color:aliceblue; }
.task { background-color:lightpink; }
</style>
```

```{r forslag, include = FALSE}
vis.forslag <- params$vis.forslag
```

```{r forklar forslag, echo = FALSE, results = "asis", include = vis.forslag}
cat('
<mark>
**LØSNINGSFORSLAG!**

Disse vises nedenfor på ulike måter:
<mark>Gulmerking</mark> av tekstforslag til løsning, eller kommentarer om alternative løsninger.
* Noe <button>Code</button> som var skjult i øvingsdokumentet.
<details>
<summary>
<mark>
Noen løsningsforslag er skjult inni hint.
</mark>
</summary>
Da må du utvide hintene for å se løsningsforslagene.  

</details>
')
```

<br> <br> <br> <br>

::: info
**Colloquium exercises:** You are encouraged to work together in groups
on the exercises. Sharing problems and solutions is often a much quicker
way to learn, especially for problems of a more practical nature.
:::

Here are some packages and data you need to use in these exercises. So
you should run these commands before you start the exercises.

```{r, echo = TRUE, message=FALSE}
library(mixlm)
library(car)
library(effects)
library(BIAS.data)
data(NSRdata)
data(Barley)
data(birth)
```

# Ex-2: Yield of barley

In `Barley` there are results from an experiment where the response is
yield of barley pr 1000 square meter, and the factors varieties of
barley, soil types, types of fertilizers. In addition was the experiment
done in two different geographical areas (sites).

-   Assume a two factor model including the main effects of `Variety`
    and `Soil` and their interaction. State the model and explain all
    parameters under a sum-to-zero parametrization.

<details>

<summary><mark> Solution </mark></summary>

The model can be written as:
$$ y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}$$
where, $\epsilon \sim N(0, \sigma^2)$

The parametrization restrictions for "sum-to-zero"" are: $$
\begin{eqnarray} 
\sum_i{\alpha_i} = 0, & \sum_j{\beta_j} = 0, & 
\sum_{i}{(\alpha\beta)_{ij}} = 0  \text{ and }
\sum_{j}{(\alpha\beta)_{ij}} = 0
\end{eqnarray}
$$

</details>

-   Fit the model in R and estimate all parameters

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
# either
mod1 <- mixlm::lm(Yield ~ Variety * Soil, data = Barley, contrasts="contr.sum")
# or
mod1 <- stats::lm(Yield ~ Variety * Soil, data = Barley, contrasts = list(x = contr.sum))
summod1 <- summary(mod1, cor=FALSE)
print(summod1)
anova(mod1)
```

From the model summaries we find the estimated parameters to be:

1.  The estimated overall mean yield: $\hat{\mu}=$ `r coef(mod1)[1]`
2.  The estimated effects of Variety 1 and 2: $\hat{\alpha}_1=$
    `r coef(mod1)[2]`, $\hat{\alpha}_2 = -\hat{\alpha}_1 =$
    `r -coef(mod1)[2]`
3.  The estimated effects of Soil 1 and 2: $\hat{\beta}_1=$
    `r coef(mod1)[3]`, $\hat{\beta}_2 = -\hat{\beta}_1 =$
    `r -coef(mod1)[3]`
4.  The estimated interaction effects of Variety and Soil:
    $\hat{\alpha\beta}_{11}=$ `r coef(mod1)[4]`,
    $\hat{\alpha\beta}_{12}=-\hat{\alpha\beta}_{11}=$
    `r -coef(mod1)[4]`,
    $\hat{\alpha\beta}_{21}=-\hat{\alpha\beta}_{11}=$
    `r -coef(mod1)[4]`,
    $\hat{\alpha\beta}_{22}=-\hat{\alpha\beta}_{21}=-\hat{\alpha\beta}_{12}=$
    `r coef(mod1)[4]`
5.  The estimated within `variety:Soil` levels unexplained variability:
    $\hat{\sigma}^2=MSE=$ `r round(summod1$sigma^2, 2)`

</details>

-   Make an interaction plot and try to conclude about the presence of
    interaction from the plot. How would you explain interaction effect
    in this example for a person with experience in agriculture, but
    minimal statistical experience?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
plot(allEffects(mod1,confidence.level = .95))
```

From the plot, it seems that a farmer can expect an increase in yield
when using Variety 2 instead of Variety 1 for Soil = Clay, but vice
versa for Soil = Sand. If this is so, there is a so-called interaction
effect of Variety and Soil on yield.

</details>

-   Perform a hypothesis test for the interaction effect. Conclusion?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
anova(mod1)
```

The interaction between `Soil` and `Variety` is insignificant at 5%
testlevel (p-value \> 0.05). This result indicates that the slight
non-parallel tendency of the lines in the interaction plot may be due to
random errors. More data would be needed to verify any interaction
effect.

</details>

-   Fit a four factor model with Variety, Soil, Fetilizer and Site as
    factors and with all possible 2-factor, 3-factor and 4-factor
    interactions. Are there any higher order significant effects?
    Compare the R-squared and the adjusted R-squared values. What can
    you conclude from these?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
mod2 <- stats::lm(Yield ~ Variety * Soil * Fertilizer * Site, data = Barley)
anova(mod2)
summary(mod2)
```

There is only one main effect and one interaction effect significant at
5% levels, namely the `Fertilizer` main effect and the `Soil:Fertilizer`
interaction. However, the least squares estimator is prone to inflated
variance estimates and difficulties in finding significance when the
number of variables `k` rises compared to the number of observations `N`
and if variables are highly correlated. Too many variables to estimate
from a limited number of observations leaves few degrees of freedom to
SSE, and tests with low power. Lack of significance may be an
over-fitting problem. By removing some of the least significant
variables, the problem is reduced, and a reduced model with significant
effects may be identified.

The large difference between $R^2$ and $R_{adj}^2$ is also an indication
of an over-fitted model.

</details>

-   Try to reduce the model by excluding the least significant higher
    order interactions first, one at a time, in a backward elimination
    procedure using the 'backward' function from the `mixlm` package.
    Use the 'hierarchy = TRUE' argument to ensure that the so-called
    marginality principle is followed. What is the final reduced model?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
mod2.red <- backward(mod2, alpha = 0.05, full = TRUE, hierarchy = TRUE)
anova(mod2.red)
```

The final reduced model is a model with all four main effects, three of
which are now significant, five 2-order interactions (three
significant), and two significant 3-order interactions. The 4-order
interaction is removed. The marginality principle says that we should
keep all main effects and interactions that are part of higher order
significant interaction(s) in the model. This is a good principle for
practical data analysis.

</details>

-   Perform a model check of the reduced model using residual analysis.

<details>

<summary><mark> Solution </mark></summary>

The `backward` function has fitted the reduced model:

```{r, echo=TRUE, fig.height=7}
par(mfrow = c(2,2))
plot(mod2.red, which = c(1,2,4,5))
```

The residual plots give no evidence of any problems with the model
assumptions.

</details>

-   You can make the third order interaction effects plot between
    `Variety`,`Site` and `Soil` like this:

```{r effectPlot, eval = TRUE, echo=TRUE}
eff <- allEffects(mod2.red, confidence.level = .95)
plot(eff[2])
```

Try to explain the plot. Why do you think the interaction between
`Variety`,`Site` and `Soil` was significant?

<details>

<summary><mark> Solution </mark></summary>

The interaction effect is significant because the lines are not
parallel. The effect of Soil on Yield is different for the two Varieties
and the effect of Soil on Yield for one Variety may depend on Site. This
is a three-way interaction effect.

</details>

# Ex-2: Indicator variable (ANCOVA model)

This exercise is a follow up on the exercise from week 3 on multiple
regression using the `mtcars` data set.

-   The `am` variable is really an indicator variable for transmission
    system of the cars (The variable is coded as 0=automatic, 1=manual).
    Fit a multiple linear model with "cyl", "wt", "am" and the
    interaction between "wt" and "am" and report the summary of the
    model with parameter estimates and tests.

NOTE: Make sure the `am` is treated as a categorical variable in the
model by writing `factor(am)` in the model formula in the linear model
function. Use the default "sum to zero" parametrization when you fit the
model.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
cars <- lm(mpg ~ cyl + wt * factor(am), data=mtcars, contrasts="contr.sum")
summary(cars)
```

</details>

-   Write up the assumed model which has been run here. Also write up
    the estimated models for automatic and manual transmission,
    respectively.

<details>

<summary><mark> Solution </mark></summary>

The model is:

$$ y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_2\cdot x_3 + \epsilon $$

where $y$=`mpg`, $x_1$=`cyl`, $x_2$=`wt`, $x_3$=`am` and
$\epsilon \sim N(0, \sigma^2)$.

For automatic transmission (`am`=$x_3$=1) we have the estimated model:
$$ 
\begin{split}
\hat{y} &= 40.2523 - 1.1814x_1 - 4.4676x_2 - 5.9693 + 2.0987x_2\\
&=34.283 - 1.1814x_1 - 2.3689x_2
\end{split}
$$

For manual transmission (`am`=$x_3$=-1) we have $$ 
\begin{split}
\hat{y} &= 40.2523 - 1.1814x_1 - 4.4676x_2 + 5.9693 - 2.0987x_2\\
&= 46.2216 - 1.1814x_1 - 6.5663x_2
\end{split}
$$

We observe that the negative effect of weight on mileage is larger for
manual transmission than for automatic.

</details>
