---
title: "Exercises STAT340 - Regression analysis - part II"
author: "Solve Sæbø" 
date: "Feb 18 2025"
params:
  vis.forslag: false
output:
  html_document:
    toc: true
    theme: yeti
    code_folding: hide
    number_sections: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style>
.main-container, .main-content { max-width: 100%; }
.info { background-color:aliceblue; }
.task { background-color:lightpink; }
</style>
```

```{r forslag, include = FALSE}
vis.forslag <- params$vis.forslag
```

```{r forklar forslag, echo = FALSE, results = "asis", include = vis.forslag}
cat('
<mark>
**LØSNINGSFORSLAG!**

Disse vises nedenfor på ulike måter:
<mark>Gulmerking</mark> av tekstforslag til løsning, eller kommentarer om alternative løsninger.
* Noe <button>Code</button> som var skjult i øvingsdokumentet.
<details>
<summary>
<mark>
Noen løsningsforslag er skjult inni hint.
</mark>
</summary>
Da må du utvide hintene for å se løsningsforslagene.  

</details>
')
```

<br> <br> <br> <br>

::: info
**Colloquium exercises:** You are encouraged to work together in groups
on the exercises. Sharing problems and solutions is often a much quicker
way to learn, especially for problems of a more practical nature.
:::

<br>

# The mtcars data (on CANVAS)

We will use an old data set from 1974 on gasoline consumption for
various cars. The data set is called `mtcars`. Load down the
mtcars.Rdata file from CANVAS and load it into RStudio.

```{r, echo=FALSE}
head(mtcars)
```

a)  Fit a multiple linear regression model with mpg (Miles/Gallon) as
    response variable and `wt` (Weight) and `cyl` (Number of cylinders)
    as predictors. Call the model object "`cars1`".

```{r, echo=FALSE}
options(contrasts=c('contr.sum','contr.poly'))
cars1 <- lm(mpg ~ cyl + wt, data=mtcars)
```

b)  Check the model assumptions by residual analysis.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=FALSE}
par(mfrow = c(2,2)) #This creates a layout for plots, one row and two columns
plot(cars1, ask=FALSE)
```

</details>

c)  Report the regression results and compute the ANOVA-table.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=FALSE}
summary(cars1)
anova(cars1)
```

</details>

d)  Write a short report on the results.

<details>

<summary><mark> Solution </mark></summary>

Both `cyl` and `wt` appears to be highly significant predictors for
`mpg`. The estimated effects are negative implying that the mileage
decreases as both weight and cylinder numbers increase, which is a
reasonable result. The $R^2$ is `r round(summary(cars1)$r.squared,3)`,
hence, about `r round(summary(cars1)$r.squared*100,0)`% of the
variability in mileage is explained by the linear relationship with
`cyl` and `wt`. The residual plot of fitted values versus residuals
gives an indication of a non-linear relationship, which may be a result
of non-linear dependencies or missing explanatory variable(s). The
normal probability plot is more or less OK.

</details>

# Influential measurements

a)  Use the influence.measures function to compute the Cook's distances
    and the leverage (hat) values for all observations according to the
    `cars1` model. A rule of thumb is that Cook's distance (column
    'cook.d') larger than 1, and leverage (column 'hat') above $2p/n$
    should be checked for influence. Are there any influential
    observations according to these measures?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=FALSE}
influence.measures(cars1)
```

Three observations have larger leverage (hat-value) than
$2p/n = 2 \times 3/32 = 0.1875$. These are potentially influential
observations. However, their Cook's distances are far below 1, hence
their positions in the predictor space do not influence the estimated
model noteworthy.

</details>

# Model selection

a)  Fit a third multiple linear regression model with mpg (Miles/Gallon)
    as response variable and `cyl`, `disp`, `hp`, `drat`, `wt` and
    `qsec` as predictor variables. Call the model object "`cars2`".
    Report a summary of the analysis.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=FALSE}
cars2 <- lm(mpg ~ cyl + disp + hp + drat + wt + qsec, data = mtcars)
summary(cars2)
```

Apparently only wt is significant, but having many variables in a model
may lead to inflated standard errors of the estimates due to correlation
between predictors, and problems finding truly significant variables.

</details>

b)  We would like to check various sub-models of this model by combining
    different variables. Use the "best subsets" method in the
    mixlm-package to find a good reduced model based on adjusted
    R-squared. Use the 'nbest=3' argument to get the three best models
    for each model size. Which model would you select?

```{r, echo=FALSE}
library(mixlm)
best.subsets(cars2, nbest=3)
```

The model size is given in the first column. Column two is the rank
within model size, then comes a column for each variable with a star
indicating that a given variable is part of the model. Finally comes the
residual sum of squares (RSS or SSE), $R^2$, $R^2$-adjusted and finally
a diagnostic called Mallow's Cp (not covered in this course). Which
sub-model would you say is the best fitting model according to the
$R^2$-adjusted?

<details>

<summary><mark> Solution </mark></summary>

A couple of models are quite similar, but the largest $R^2$-adjusted is
obtained with a model with predictors `cyl`, `hp` and `wt`. This is also
a quite simple model with few predictors. We should always strive for
simple models and choose the simpler model in cases where the fit
appears to be more or less equal for several models.

</details>

## Optional mathematical exercise

The last model you fitted last week to Sales and TV (Advertising data)
on the original scale was really:
$$y_i = \beta_0\cdot x_i^{\beta_1}\cdot \epsilon_i$$ which is a
so-called multiplicative model. Show that the transformations you made
to Sales and TV in exercise i. transforms this model into a linear
model. How do the estimated regression coefficients of the model you
fitted in i. relate to the $\beta_0$ and the $\beta_1$ in the
multiplicative model?

<details>

<summary><mark> Solution </mark></summary>

Taking the logarithm on both sides we obtain:

$$
\begin{split}
log(y_i) &= log(\beta_0\cdot x_i^{\beta_1}\cdot \epsilon_i))\\
log(y_i) &= log(\beta_0) + log(x_i^{\beta_1}) + \log(\epsilon_i)\\
log(y_i) &= log(\beta_0) + \beta_1log(x_i) + \log(\epsilon_i)\\
y_i^* &= \beta_0^* +\beta_1 x_i^* + \epsilon_i^*
\end{split}
$$ where $y_i^*$ is logSales, $x_i^*$ is logTV. We can see that the
intercept we estimated in the transformed model
($\hat{\beta}_0^*=0.905$) really is the logarithm of the $\beta_0$ in
the multiplicative model. Hence we can find
$\hat{\beta}_0 = \exp(\hat{\beta_0^*})=exp(0.905) = 2.47$. The $\beta_1$
is un-transformed, hence $\hat{\beta}_1 = 0.355$. Furthermore, the error
terms in the transformed model are the logarithms of the error terms of
the multiplicative model.

To verify this we can add the estimated multiplicative model to a
scatterplot of Sales and TV-budget, as in this code:

```{r, echo=FALSE, message=FALSE, results='hide'}
load("Advertising.Rdata")
plot(Sales~TV, pch=20, cex=2, col="red", data=Advertising)
curve(2.47*x^(0.355), add=TRUE, lwd=2, col="blue")
```

</details>
