---
title: "Exercises STAT340 - GLM 2"
params:
  vis.forslag: false
output:
  html_document:
    toc: true
    theme: yeti
    code_folding: hide
    number_sections: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style>
/* Disable interaction with all details/summary */
details summary {
  pointer-events: none;   /* no mouse clicks */
  user-select: none;      /* cannot select text */
}

/* Optionally hide the disclosure triangle marker */
details summary::marker {
  content: "";
}
</style>
```

```{=html}
<style>
.main-container, .main-content { max-width: 100%; }
.info { background-color:aliceblue; }
.task { background-color:lightpink; }
</style>
```

```{r forslag, include = FALSE}
vis.forslag <- params$vis.forslag
```

```{r forklar forslag, echo = FALSE, results = "asis", include = vis.forslag}
cat('
<mark>
**LØSNINGSFORSLAG!**

Disse vises nedenfor på ulike måter:
<mark>Gulmerking</mark> av tekstforslag til løsning, eller kommentarer om alternative løsninger.
* Noe <button>Code</button> som var skjult i øvingsdokumentet.
<details>
<summary>
<mark>
Noen løsningsforslag er skjult inni hint.
</mark>
</summary>
Da må du utvide hintene for å se løsningsforslagene.  

</details>
')
```

<br> <br> <br>


You will need these libraries and files for these exercises:

```{r, echo = TRUE, message=FALSE}
library(BIAS.data)
library(nnet)
load("taxonomy.Rdata")
data(species)
source("summaryMultinom.R") # Can be downloaded from the CANVAS module.
```

# Ex-1: Count data

Consider the plant species count data `species.Rdata`.

a.  Redo the analysis from the lesson using the `Biomass` and `pH`
    variables as predictors for species counts in a Poisson regression
    model. Note that the default parametrization for GLM is "reference
    level" which here means that the effect of PH=High is set to zero.
    Find separate expressions for the estimated linear predictors
    $\widehat{\eta}_j$ for each $j=$ `pH`-levels =
    "`low`","`medium`","`high`". Also find the estimated expected count
    for `Biomass = 2` and `pH = high`.

<details>

<summary><mark> Solution </mark></summary>

First we fit the model:

```{r, echo=FALSE}
options(contrasts=c("contr.treatment", "contr.poly"))
mod2.glm <- glm(Species ~ Biomass + pH, family=poisson(log), data=species)
summary(mod2.glm)
```

Then the estimated linear predictors are:

$$\eta_\text{low} = (3.85 - 1.14) - 0.128\cdot \text{Biomass} = 2.71 - 0.128\cdot \text{Biomass}$$
$$\eta_\text{medium} = (3.85 - 0.45) - 0.128\cdot \text{Biomass} = 3.4 - 0.128\cdot \text{Biomass}$$
$$\eta_\text{high} = 3.85 - 0.128\cdot \text{Biomass}$$ For
`Biomass = 2` and `pH = high` the estimated expected count is:

$$\widehat{\lambda} = \exp(3.85 - 0.128\cdot 2) = 36.4$$

```{r, echo=FALSE}
    link <- predict(mod2.glm, data.frame('Biomass'=2, 'pH'='high'), type = "link", se.fit = TRUE)
    response <- predict(mod2.glm, data.frame('Biomass'=2, 'pH'='high'), type = "response", se.fit = TRUE)
    LS <- cbind(link$fit, link$se.fit, response$fit, response$se.fit)
    colnames(LS) <- c("link", "se.link", "response", "se.response")
    LS
```

If we compute the estimated expected counts (predicted values) for
different values of Biomass for each of the three pH-levels in a similar
way, we may obtain fitted curves as shown below. Verify from the plot
that Biomass=2 for pH='high' gives approximately a fitted value of 36.4.

```{r,echo=FALSE}
attach(species)
plot(Biomass, Species, type = "n")
spp <- split(Species, pH)
bio <- split(Biomass, pH)
points(bio[[1]], spp[[1]], pch = 16, col = "red")
points(bio[[2]], spp[[2]], pch = 16, col = "green")
points(bio[[3]], spp[[3]], pch = 16, col = "blue")

legend(x = "topright", legend = c("high", "medium", "low"), 
       pch = 16, col = c("red", "blue", "green"), title = "pH")

curve(exp(3.85 - 0.128 * x), add = T, lwd = 2, col = "red")
curve(exp(3.4 - 0.128 * x), 
      add = T, lwd = 2, col = "blue")
curve(exp(2.71 - 0.128 * x), 
      add = T, lwd = 2, col = "green")
detach(species)
```

</details>

b.  Fit the same model as in a. using the quasipoisson approach. Is
    there any evidence of overdisperion?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=FALSE}
mod2q.glm <- glm(Species ~ Biomass + pH, family = quasipoisson, data = species)
summary(mod2q.glm)
```

The dispersion parameter is only marginally larger than 1. We conclude
that over-disperion is not a big problem here.

</details>

c.  Extend the model from a. by including an interaction between Biomass
    and pH. Test the significance of the interaction. Is there a
    significant difference between pH="high" and pH="medium" with regard
    to the effect of Biomass to the species count?

<details>

<summary><mark> Solution </mark></summary>

First we fit the model with interaction term and get these results:

```{r, echo=FALSE}
mod3.glm <- glm(Species ~ Biomass * pH, family=poisson(log), data=species)
summary(mod3.glm)
```

The approximate normal tests (the z-test/Wald tests) only give
significance tests for whether individual parameters are equal to zero.
For instance, the Wald test gave a p-value of 0.167 for the term
`Biomass:pH(mid)` which really is the contrast between interactions
`Biomass:pH(high)`and `Biomass:pH(mid)` since `pH(high)` is the
reference level in this model. The relatively large p-value says that
there is no evidence that Biomass has different effect on counts for
these two pH-levels. On the other hand there appears to be a significant
difference in the effect of Biomass between `pH(high)` and `pH(low)`
with a p-value of 0.000108. In order to evaluate the *overall*
significance of the interaction term (Biomass:pH) we can look at the
deviance tests in the Analysis of Deviance Table:

```{r, echo=FALSE}
anova(mod3.glm,test='LR')
logLik(mod3.glm)
```

Here we can read out that adding the interaction term to a model already
containing Biomass and pH will reduce the deviance with 16.040 and the
qhi-square test on this deviance with 2 degrees of freedom returns a
p-value of 0.00033, hence a highly significant interaction.

We could also get the same result by comparing the interaction model (as
a full model) with the model from exercises a) (as a reduced model) and
then use Models \> Hypothesis test \> Compare two models

```{r, echo=FALSE}
anova(mod2.glm, mod3.glm, test="Chisq")
```

The interaction term is again highly significant with a p-value of
0.00033.

A corresponding plot as made in a) including interaction effect yields:

```{r,echo=FALSE}
attach(species)
plot(Biomass, Species, type = "n")
spp <- split(Species, pH)
bio <- split(Biomass, pH)
points(bio[[1]], spp[[1]], pch = 16, col = "red")
points(bio[[2]], spp[[2]], pch = 16, col = "green")
points(bio[[3]], spp[[3]], pch = 16, col = "blue")

legend(x = "topright", legend = c("high", "medium", "low"), 
       pch = 16, col = c("red", "blue", "green"), title = "pH")

curve(exp(3.76812 - 0.10713 * x), add = T, lwd = 2, col = "red")
curve(exp(3.76812 - 0.33146 - (0.10713 + 0.03189) * x), 
      add = T, lwd = 2, col = "blue")
curve(exp(3.76812 - 0.81557 - (0.10713 + 0.155028) * x), 
      add = T, lwd = 2, col = "green")
detach(species)
```

Note now how the curves differ from those found in a). That change is
the interaction effect.

</details>

# Ex-2: Multinomial data

We return to the taxonomy data used in the lecture on cluster analysis.
In the file `taxonomy.Rdata` you find seven variables measured on plants
from four different taxa.

a.  Fit a multinomial regression with `Taxon` as the multinomial
    categorical response variable and `Sepal` as predictor. Is the
    effect of `Sepal` on the probabilities of the various taxons
    sifnificant at a 5% test level? Download the "summaryMultinom.R"
    file from the CANVAS-module and source it in R. This file contains
    code to summarize the multinomial regression model through the
    summaryMultinom() function which can be applied to the fitted
    multinomial model. The function will give you the estimated
    regression coefficients, the standard errors, the z-values and the
    p-values for each of the regression coefficients.

<details>

<summary><mark> Solution </mark></summary>

```{r}
library(nnet, pos=22)
source("summaryMultinom.R")
mod1.multi <- multinom(Taxon ~ Sepal, data=taxonomy, trace=FALSE)
summaryMultinom(mod1.multi)
```

The Deviance test gives a p-value much smaller than 0.05, and `Sepal` is
found to influence the probability of taxons significantly.

</details>

b.  What are the estimated linear predictors $\widehat{\eta}_j$ for each
    of the four taxa, $j=1,2,3,4$?

<details>

<summary><mark> Solution </mark></summary>

First we extract the estimated regression coefficients. Remember that
for the reference level taxon (`taxon=1`) the regression coefficients
are assumed equal to zero. We may sum up the estimated coefficients like
this:

```{r}
summod1.multi <- summary(mod1.multi)
co <- summod1.multi$coefficients
co <- rbind(c(0,0),co)
rownames(co)[1] <- "I"
co
```

This gives the following linear predictors:
$$\widehat{\eta}_1 = 0 + 0\cdot \text{Sepal} = 0$$
$$\widehat{\eta}_2 = 1.49 - 0.593\cdot \text{Sepal}$$
$$\widehat{\eta}_3 = 2.86 - 1.15\cdot \text{Sepal}$$
$$\widehat{\eta}_4 = -48.0 + 13.7\cdot \text{Sepal}$$

</details>

c.  What is the most probable taxon according to this model for a plant
    with `Sepal=3`? (Hint: First insert `Sepal=3` in the linear
    predictors you just found to obtain the actual value of this. Then
    use the expression for the probability of a outcome $j$:

$$\widehat{p}_j=\frac{e^{\widehat{\eta}_j}}{\sum_k e^{\widehat{\eta}_k}}= \frac{e^{\widehat{\eta}_j}}{ e^{\widehat{\eta}_1}+e^{\widehat{\eta}_2}+e^{\widehat{\eta}_3}+e^{\widehat{\eta}_4}}$$

<details>

<summary><mark> Solution </mark></summary>

This should give the estimated probabilities:

```{r, echo=FALSE}
x <- 3
probs <- exp(co[ ,1] + co[ ,2] * x) / sum(exp(co[ ,1] + co[ ,2] * x))
probs
```

We see that taxon I is the most probable with a probability equal to
`0.43`.

Extra: The following figure shows the probabilities of each taxon for
various values of `Sepal`. Remember that these probabilities are the
posterior probability estimates for the class in a classification
terminology. We classify to the most probable class given the observed
predictor values.

```{r, echo=FALSE}
xvector <- seq(0,7, by = 0.05)
probfunc <- function(x){
  exp(co[ ,1] + co[ ,2] * x) / sum(exp(co[ ,1] + co[ ,2] * x))
  }
probmat <- t(sapply(xvector, probfunc))
matplot(xvector, probmat, lty = 1, type = "l", lwd = 2, 
        main = "Posterior probabilities for taxa", 
        xlab = "Sepal", ylab = "Estimated posterior probability")
legend(6, 0.8, legend = c("I", "II", "III", "IV"), lty = 1, col = 1:4, lwd = 2)
```

We observe that taxon III is the most probable for `Sepal<2.5`, taxon I
for `2.5 < Sepal < 3.5` and taxon IV for `Sepal >= 3.5`. Taxon 2 is the
most probable taxon only for `Sepal` $\approx$ `2.5`.

</details>

d.  Create a Boxplot of `Sepal` from the Graph menu and choose to plot
    by groups identified by the taxon levels. Can you see why Sepal
    appears to be informative with regard to taxon?

<details>

<summary><mark> Solution </mark></summary>

```{r}
plot(Sepal ~ Taxon, data = taxonomy)
```

We see that taxon 4 has quite different Sepal values than the other
taxons, hence Sepal is a good variable to separate taxon 4 from the
other 3 taxons.

</details>

e.  Use similar plots like in d. to identify other variables which are
    promising with regard to separating taxa I, II and III (Sepal is
    already a good variable to distinguish taxon IV from the others).
    Fit an extended model with the variables you have chosen.

<details>

<summary><mark> Solution </mark></summary>

I have found two variables which are promising; `Leaf` and `Petiole`, as
shown in the figures below.

```{r, echo=FALSE}
plot(Leaf ~ Taxon, data = taxonomy)
plot(Petiole ~ Taxon, data = taxonomy)
```

`Leaf` seems to distinguish taxon III from the others, and `Petiole` may
separate taxa I and II.

I therefore fit a model with `Sepal`, `Leaf` and `Petiole` as predictors
for taxa:

```{r}
mod2.multi <- multinom(Taxon ~ Sepal + Leaf + Petiole, data=taxonomy, trace=FALSE)
summaryMultinom(mod2.multi)
```

</details>
