---
title: "Exercises STAT340 - PCA"
params:
  vis.forslag: false
output:
  html_document:
    toc: true
    theme: yeti
    code_folding: hide
    number_sections: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style>
/* Disable interaction with all details/summary */
details summary {
  pointer-events: none;   /* no mouse clicks */
  user-select: none;      /* cannot select text */
}

/* Optionally hide the disclosure triangle marker */
details summary::marker {
  content: "";
}
</style>
```

```{=html}
<style>
.main-container, .main-content { max-width: 100%; }
.info { background-color:aliceblue; }
.task { background-color:lightpink; }
</style>
```

```{r forslag, include = FALSE}
vis.forslag <- params$vis.forslag
```

```{r forklar forslag, echo = FALSE, results = "asis", include = vis.forslag}
cat('
<mark>
**LØSNINGSFORSLAG!**

Disse vises nedenfor på ulike måter:
<mark>Gulmerking</mark> av tekstforslag til løsning, eller kommentarer om alternative løsninger.
* Noe <button>Code</button> som var skjult i øvingsdokumentet.
<details>
<summary>
<mark>
Noen løsningsforslag er skjult inni hint.
</mark>
</summary>
Da må du utvide hintene for å se løsningsforslagene.  

</details>
')
```

<br> <br> <br>

# Track and Field data

The data set `trackfieldrecords` is a `list` with two data tables: The
`trackfieldrecords$runMen` and `trackfieldrecords$runWomen` containing
national records (a few years ago…) for several track and field events
like 100m, 200m and so on up to marathon. <br>

## Join data (data wrangling)

Instead of having separate data for men and women, let us join the data
into a single table. This means we *join* the data tables by putting
them next to each other, but making certain the rows (countries) are
aligned correctly. First we load the data and inspect the two data
tables

```{r, results='hide', message=F}
library(tidyverse)
library(BIAS.data)

data(trackfieldrecords)
head(trackfieldrecords$runMen)  # men's records
head(trackfieldrecords$runWomen)  # women's records
```

We notice that the womens records have an `M` before each distance, just
like the men. This must be wrong? Let us replace the `M` by a `W`, to
indicate which records are for men and which for women:

```{r}
old.names <- colnames(trackfieldrecords$runWomen)
new.names <- str_replace(old.names, "M", "W")
colnames(trackfieldrecords$runWomen) <- new.names
```

Then we are ready to join the two tables, and we use the column `Nation`
as the joining information. This will ensure all records for the same
nation are in the same row. This is important, it may be that nations
are listed in different order in the two tables. Never ever assume such
information is listed in the exact same order in two different data
tables!

```{r}
all.records <- full_join(trackfieldrecords$runMen, trackfieldrecords$runWomen, by = "Nation") %>% 
  select(Nation, starts_with("W"), starts_with("M"))  # just a new ordering of the columns
```

<br>

## Inspect data

Explore the data by making a correlation plot. for example by using
`ggcorr()`-function from the `GGally`-package (you need to install it).

```{r, message=F}
library(GGally)
ggcorr(all.records)
```

What is the big picture we see from this?

<details>

<summary><mark> Solution </mark></summary>

All correlations are positive, and also pretty high. These are
correlations between the variables. It simply means that if a nation has
good/bad running records for one variable, it has similar good/bad for
the others as well. This indicates to us that these variables carry to a
large degree the same information. We then expect a PCA should show most
information is captured with 1 or at least very few components.

</details>

<br>

## PCA

Run a PCA with the `prcomp()`-function (and with `scale = FALSE`) on the
data and print out a `summary()` of the resulting object (do not include
the first column with names of nations). How many components is needed
to explain at least 99% of the variation in the data?

Also, print the loadings for the first two principal components. Try to
give an interpretation of them, and explain why so much of the total
variability is found in the first component.

<details>

<summary><mark> Solution </mark></summary>

```{r}
pca.obj <- prcomp(all.records[,-1], scale=FALSE)
summary(pca.obj)
pca.obj$rotation[,1:2]
```

Variation in data is now mainly due to the varying distances. Large
values for large distances and small values for short distances. This is
an example of data where PC1 picks up the 'obvious' and overwhelming
cause of why the data varies. From the loadings, see that the loading
weight for the 42k's (Marathon) is totally dominating with a weight of
0.97 (Women) and 0.23 (Men). All other weights are small. PC1 is
therefore more or less identical to the Marathon variables. Note that
this 'signal' may not be very interesting, everyone knows it takes more
time to run a Marathon than a 100m, and therefore the variance for these
variables are much bigger. We should consider scaling these data, to get
rid of this rather obvious, and uninteresting, source of variation.

</details>

<br>

## Standardizing

Re-run the PCA with option `scale. = TRUE` in `prcomp()`, which means
the variables are standardized prior to the PCA. How many variables are
needed to explain 99% of the variation in this case? How much is
explained by the two first components? How would you interpret the
loadings of PC1, PC2 and PC3?

<details>

<summary><mark> Solution </mark></summary>

```{r}
pca.obj <- prcomp(all.records[,-1], scale. = TRUE)
summary(pca.obj)
pca.obj$rotation[,1:3]
```

We need 11 components to achieve 99% explained variance.

We notice PC1 has almost the same coordinate for all variables,
indicating an average over all distances. We may think of this as a
general 'goodness' in the nation's records, regardless of distance. The
distance-effect is now in PC2 instead. In PC3 we see the effect of
gender (men versus women).

</details>

<br>

## Scores and loadings

Make a score-plot and a loading-plot with the first two components using
the code below. You may use `geom_text()` instead of `geom_point()` and
use `Nation` as the label in `geom_text()`.

```{r}
score.tbl <- as.data.frame(pca.obj$x) %>% 
  bind_cols(all.records)
relative.evar <- round(pca.obj$sdev^2 / sum(pca.obj$sdev^2), 2)
ggplot(score.tbl) +
  geom_text(aes(x = PC1, y = PC2, label = Nation)) +
  labs(x = str_c("PC1 (", relative.evar[1], ")"),
       y = str_c("PC2 (", relative.evar[2], ")"))

loading.tbl <- as_tibble(pca.obj$rotation, rownames = "Variables")
origin.tbl <- as_tibble(pca.obj$rotation - pca.obj$rotation, rownames = "Variables") %>% 
  bind_rows(loading.tbl)
ggplot(loading.tbl, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(aes(label = Variables), nudge_y = 0.05) +
  geom_line(aes(group = Variables), data = origin.tbl, color = "maroon2")
```

Give comments to how the nations distribute relative to the loadings.

Then, edit the code to plot instead PC1 versus PC3. Comment as above.

<details>

<summary><mark> Solution </mark></summary>

Consider first the score-plot with PC1-PC2 and where the countries are
located here. In such plots we typically look for either
clusters/groupings in the data, or for gradients, i.e. the objects
(countries) are 'ordered' in some direction. For those interested in
track-and-fields will notice countries with small (negative) values
along PC1 are those who tend to dominate these sports, while on the
other end we find countries rarely seen on the Olympic podium. Thus,
here PC1 simply reflects the 'goodness' of the nations in terms of their
records. A larger PC1 loading means a larger (=poorer) record.

Considering the PC1-PC2 loading-plot, i.e. the variables, we see the
clear gradient along PC2 reflecting the distances. Nations are
distributed along this axis depending if their athletes tend to be good
in short or long distances.

For the PC1-PC3 plots, the PC1 axis is as before. Here we see that the
PC3 axis separate men and women records, and mostly in the shorter
distances. Nations at the top (large PC3 score) have men's records with
larger times compared to the women's record, i.e. the records are more
equal between the genders.

</details>

<br> <br> <br>

# Sediments data

Virtually all environments on this planet may be characterized by their
*microbial composition*, i.e. which micro-organisms they inhabit and at
which abundances we observe the various taxa. This is done by
*sequencing*, and the *read-count* from sequencing is a measure of the
abundance for the various taxa.

The data `sediments` contains the read-counts of 4112 prokaryotes
(micro-organisms, mainly bacteria) in seafloor sediments collected at 12
selected locations along the Norwegian coast. Each variable is the
abundance of a named genus. A large value means a large abundance, but
the *sequencing depth* varies between the samples (rows), i.e. if you
sum the values in each row they are not identical.

The first column is the type of environment of the sample, here
described as either `Natural` or `Polluted`.

<br>

## PCA

Make a PCA of the numerical data (not `Environment`). How much of the
total variation is explained by the first PCs?

<details>

<summary><mark> Solution </mark></summary>

```{r, message=F}
library(tidyverse)
library(BIAS.data)
data(sediments)

pca.obj <- prcomp(sediments[,-1], scale. = F)
summary(pca.obj)
```

Here 98.2% of the variation is described by the first two PC's, and
almost 95% by the first PC alone. This means we essentially only need 1
principal component to describe almost all variation in these data.

</details>

<br>

## Scoreplot

Make a score-plot, and color by the `Environment` using the code below.
Imagine you are to explain this plot to someone else, what would you
say? What is your interpretation of PC1 here?

```{r, message=F}
library(tidyverse)
score.tbl <- as.data.frame(pca.obj$x) %>% 
  bind_cols(sediments)
relative.evar <- round(pca.obj$sdev^2 / sum(pca.obj$sdev^2), 2)
ggplot(score.tbl) +
  geom_point(aes(x = PC1, y = PC2, color = Environment), size = 3) +
  labs(x = str_c("PC1 (", relative.evar[1], ")"),
       y = str_c("PC2 (", relative.evar[2], ")"))
```

<details>

<summary><mark> Solution </mark></summary>

The score plot gives us the original data in the new coordinate system.
We take a look at where the points are located and look for
clusters/grouping. When colored by environment the score plot tells us
that there is a difference between natural and polluted environments.

Here the variation of the PC1 axis explains almost everything (95%), and
any variation in the y-axis direction is probably just random noise. A
plot like this should be wide but short (low), i.e. a long x-axis and a
very short y-axis, to avoid the viewer from starting to 'interpret'
variation along the y-axis.

</details>

<br>

## Loading-plot

Make a loading-plot using the code below. Again, try to explain what
this plot shows.

If we wanted to quickly determine if seafloor sediments are polluted or
not, which organisms should we look for in our samples? Is a large
abundance of this an indication of a polluted or clean environment?

```{r, message=F}
loading.tbl <- as_tibble(pca.obj$rotation, rownames = "Variables")
origin.tbl <- as_tibble(pca.obj$rotation - pca.obj$rotation, rownames = "Variables") %>% 
  bind_rows(loading.tbl)
ggplot(loading.tbl, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(aes(label = Variables), nudge_y = 0.05) +
  geom_line(aes(group = Variables), data = origin.tbl, color = "maroon2")
```

<details>

<summary><mark> Solution </mark></summary>

The loading-plot shows how strongly each variable influences a PC. The
angles between the vectors tell us how variables correlate with one
another (described above) if two variables form a small angle, the two
variables are correlated. It is hard to tell from our plot as they are
very clustered. But we can see Lutibacter and Oceanispirochaeta are
positively correlated, while Lutibacter is negatively correlated to
Viruses.

The loading-plot can tell us which variables are mostly correlated with
PC1 and PC2. We saw from the score-plot that PC1 separate polluted from
natural sites (see score plot). In the loading-plot we see that
Sulfurovum is the genus that has the (by far!) largest absolute loading
along PC1. A high abundance of this is associated with polluted sites,
since it has a negative loading along PC1, and polluted sites also had a
negative score along PC1. Thus, if we observe a very high abundance of
this taxon compared to the rest, we predict it is a polluted
environment.

</details>

<br>

## Standardizing

Try to re-run the analyses above, but standardize the data prior to the
PCA. What does these results tell us?

<details>

<summary><mark> Solution </mark></summary>

```{r}
data(sediments)

pca.obj <- prcomp(sediments[,-1], scale. = T)
summary(pca.obj)
```

When scaled, 80% of the variation is described by the first two PC's,
much less than before.

```{r}
score.tbl <- as.data.frame(pca.obj$x) %>% 
  bind_cols(sediments)
relative.evar <- round(pca.obj$sdev^2 / sum(pca.obj$sdev^2), 2)
ggplot(score.tbl) +
  geom_point(aes(x = PC1, y = PC2, color = Environment), size = 3) +
  labs(x = str_c("PC1 (", relative.evar[1], ")"),
       y = str_c("PC2 (", relative.evar[2], ")"))
```

When scaled the major difference. along PC1, is still between polluted
and natural sites, but not as clear as before.

```{r}
loading.tbl <- as_tibble(pca.obj$rotation, rownames = "Variables")
origin.tbl <- as_tibble(pca.obj$rotation - pca.obj$rotation, rownames = "Variables") %>% 
  bind_rows(loading.tbl)
ggplot(loading.tbl, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(aes(label = Variables), nudge_y = 0.05) +
  geom_line(aes(group = Variables), data = origin.tbl, color = "maroon2")
```

Well, a loading plot with this many standardized variables quickly
becomes a mess! Since all variables now have standard deviation 1.0,
they tend to be on a circle/ellipsoide around the origin. In such cases
a loading-plot is of little help to explore our data!

In fact, for data like this it makes no sense to standardize. The point
is that all variables are measured on the same scale (read-counts), and
the information lies mostly in that some organisms are more or less
abundant, i.e. have larger or smaller values. By standardizing we
eliminate much of this information! It is almost remarkable that the
score-plot still separates polluted from natural environments.

</details>

<br>
