---
title: "Exercises STAT340 - Cluster analysis"
date: "Mar 10 2026"
params:
  vis.forslag: false
output:
  html_document:
    toc: true
    theme: yeti
    code_folding: hide
    number_sections: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style>
/* Disable interaction with all details/summary */
details summary {
  pointer-events: none;   /* no mouse clicks */
  user-select: none;      /* cannot select text */
}

/* Optionally hide the disclosure triangle marker */
details summary::marker {
  content: "";
}
</style>
```

```{=html}
<style>
.main-container, .main-content { max-width: 100%; }
.info { background-color:aliceblue; }
.task { background-color:lightpink; }
</style>
```

```{r forslag, include = FALSE}
vis.forslag <- params$vis.forslag
```

```{r forklar forslag, echo = FALSE, results = "asis", include = vis.forslag}
cat('
<mark>
**LØSNINGSFORSLAG!**

Disse vises nedenfor på ulike måter:
<mark>Gulmerking</mark> av tekstforslag til løsning, eller kommentarer om alternative løsninger.
* Noe <button>Code</button> som var skjult i øvingsdokumentet.
<details>
<summary>
<mark>
Noen løsningsforslag er skjult inni hint.
</mark>
</summary>
Da må du utvide hintene for å se løsningsforslagene.  

</details>
')
```

<br> <br> <br>

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE)
library(tidyverse)
library(ggdendro)
library(knitr)
library(BIAS.data)
data(Birds)
#load("C:\\Users\\idsaksei\\OneDrive - Norwegian University of Life #Sciences\\STAT340 V24\\data\\WinterBirds4.RData")
#load("C:\\Users\\idsaksei\\OneDrive - Norwegian University of Life #Sciences\\STAT340 V24\\data\\WinterBirds.RData")
```

Make sure you have installed and loaded the following packages before
you start with today's exercises.

```{r libraries, eval = FALSE}
library(tidyverse)
library(ggdendro)
library(ggdendro)
library(BIAS.data)
```

## Exercise 1 - Bird presence & abundance in Norway

In the figure below, the observed mid-winter abundances of four
different birds species have been marked with red colour (source:
Jørgensen & Svorkmo-Lundberg (2006) Norsk vinterfuglatlas). Birds were
observed in 10 km^2^ squares accross entire mainland Norway, which was
divided into 3163 squares in total. Every square is considered one
variable. In this exercise, we will consider the 25 most abundant winter
bird species in Norway, starting with the four displayed in the figure.
In this exercise, we want to cluster birds into groups based on their
geographical distribution during winter using hierarchical clustering
methods. Which birds have similar distribution in Norway during winter?

<center>

![Fig. 1: Abundance of four common bird species in Norway during
winter.](WinterBirds4.png){#id .class width="75%" height="75%"}

</center>

We will use data containing maximum abundance measures of each of 25
birds in the 3163 squares. The value reflects the largest simultaneous
count of a given bird species in a square region.

a)  Below there are three tables. In the table on top are the Euclidean
    distances between the abundance measures for the four species, A--D,
    shown in the figure above. The second table is the table after
    merging species C and D into the first cluster (called C/D). Use the
    single linkage method to fill in the missing distances in this
    table. Finally, there is a 2 × 2 table at the bottom for the second
    clustering iteration. Write down the table and fill in the missing
    "cluster names" (?/?) as well as the missing distances (`NA`).

    Before the 1^st^ iteration:

```{r iter1, eval = TRUE, echo = FALSE, indent = "   ", fig.align = "center", rows.print = 5}
d <- matrix(c(0, 125.2, 129.0, 126.3, 125.2, 0, 106.7, 126.3, 129.0, 106.7, 0, 96.7, 126.3, 126.3, 96.7, 0), nrow = 4, ncol = 4, byrow = TRUE)
rownames(d) <- c("A", "B", "C", "D")
colnames(d) <- rownames(d)
knitr::kable(d)
```

After the 1^st^ iteration:

```{r iter2, eval = TRUE, echo = FALSE, indent = "   ", fig.align = "center", rows.print = 5}
d <- matrix(c(rep(NA, 9)), nrow = 3, ncol = 3)
rownames(d) <- c("A", "B", "C/D")
colnames(d) <- rownames(d)
knitr::kable(d)
```

After the 2^nd^ iteration:

```{r iter3, eval = TRUE, echo = FALSE, indent = "   ", fig.align = "center", rows.print = 5}
d <- matrix(rep(NA, 4), nrow = 2, ncol = 2)
rownames(d) <- rep("?/?", 2)
colnames(d) <- rownames(d)
knitr::kable(d)
```

<details>

<summary><mark> Solution </mark></summary>

After the 1^st^ iteration:

```{r iter2sol, eval = TRUE, echo = FALSE, indent = "   ", fig.align = "center", rows.print = 5}
d <- matrix(c(0, 125.2, 126.3, 125.2, 0, 106.7, 126.3, 106.7, 0), nrow = 3, ncol = 3)
rownames(d) <- c("A", "B", "C/D")
colnames(d) <- rownames(d)
knitr::kable(d)
```

After the 2^nd^ iteration:

```{r iter3sol, eval = TRUE, echo = FALSE, indent = "   ", fig.align = "center", rows.print = 5}
d <- matrix(c(0, 125.2, 125.2, 0), nrow = 2, ncol = 2)
rownames(d) <- c("A", "B/C/D")
colnames(d) <- rownames(d)
knitr::kable(d)
```

</details>

b)  Sketch the corresponding dendrogram for this hierachical clustering.

<details>

<summary><mark> Solution </mark></summary>

```{r dendrogram, fig.align='center'}
d <- matrix(c(0, 125.2, 129.0, 126.3, 125.2, 0, 106.7, 126.3, 129.0, 106.7, 0, 96.7, 
              126.3, 126.3, 96.7, 0), nrow = 4, ncol = 4, byrow = TRUE)
rownames(d) <- c("A", "B", "C", "D")
colnames(d) <- rownames(d)

clust <- hclust(as.dist(d), method = "single")
plot(clust, hang = -1, main = "Single linkage",xlab = "", sub = "")
```

</details>

c)  Load the data for the four most common birds using `data(Birds4)`
    and run hierarchical clustering in RStudio with single linkage and
    Euclidean distances. Plot the dendrogram. This should yield the same
    dendrogram as the one you sketched in exercise b). Use the results
    from here to check your answer in exercise a).

<details>

<summary><mark> Solution </mark></summary>

```{r hclust1, eval = TRUE, echo = TRUE, indent = "   ", fig.align = "center", rows.print = 5}
data(Birds4)

edistBirds4 <- dist(Birds4, method = "euclidean")     # calculating an Euclidean distance matrix
clustBirds4 <- hclust(edistBirds4, method = "single") # hierachical clustering with single linkage

# Plot dendrogram using the base R package
plot(clustBirds4, hang = -1, main = "Single linkage", xlab = "", sub = "")

# Plot dendrogram using the ggdendro package
ggdendrogram(clustBirds4, rotate = TRUE)
```

-   The distance matrix `round(edistBirds4, digits = 1)` is equivalent
    to the table that you started with (before the first iteration).

-   The distances between the clusters are stored in the `height`
    element of the dendrogram. You can retrieve it with the command
    `round(clustBirds4$height, digits = 1)`.

Here's how to check your answers in a):

```{r, include=FALSE}
data(Birds4)
edistBirds4 <- dist(Birds4, method = "euclidean")     # calculating an Euclidean distance matrix
clustBirds4 <- hclust(edistBirds4, method = "single") # hierachical clustering with single linkage
```

```{r}
round(clustBirds4$height, digits = 1)
```

-   The 1^st^ number in the distance between the groups that were merged
    into one cluster during the first iteration. Those were the species
    with the *smallest* Euclidean distance, mallard (C) and crow (D).
-   The 2^nd^ number is the distance between the newly formed cluster
    (C, D) and the species that is closest to one of the elements in (C,
    D)  -- remember that we use single linkage here! The shortest
        distance is (C, D)--B = 106.7.
-   The 3^rd^ value is the shortest distance between the cluster formed
    by C, D, & B and the last species (A), i.e. ((C, D), B)--A = 125.2.

</details>

d)  Load the file `data(Birds)` containing data for 25 bird species.
    Compute a Euclidean distance matrix and compare different linkage
    methods with dendrograms.

<details>

<summary><mark> Solution </mark></summary>

Creating the distance matrix with `dist`

```{r dist2, eval = TRUE, echo = TRUE, indent = "   ", fig.align = "center"}
edistBirds <- dist(Birds, method = "euclidean")
```

We will plot with `ggdendro`, which allows us to rotate the dendrogram
and thus make it more reader-friendly.

Single linkage:

```{r hclust2a, eval = TRUE, echo = TRUE, indent = "   ", fig.align = "center"}
clustBirds.sin <- hclust(edistBirds, method = "single")

ggdendrogram(clustBirds.sin, rotate = TRUE) + ggtitle("Single linkage")
```

Complete linkage:

```{r hclust2b, eval = TRUE, echo = TRUE, indent = "   ", fig.align = "center"}
clustBirds.com <- hclust(edistBirds, method = "complete")

ggdendrogram(clustBirds.com, rotate = TRUE) + ggtitle("Complete linkage")
```

Average linkage:

```{r hclust2c, eval = TRUE, echo = TRUE, indent = "   ", fig.align = "center"}
clustBirds.avg <- hclust(edistBirds, method = "average")

ggdendrogram(clustBirds.avg, rotate = TRUE) + ggtitle("Average linkage")
```

</details>

e)  Repeat exercise d) with average linkage, but change the distance
    method from Euclidean to Manhattan. Compare the two average linkage
    dendrograms. Does the grouping change much?

<details>

<summary><mark> Solution </mark></summary>

```{r hclust3, eval = TRUE, echo = TRUE, indent = "   ", fig.align = "center"}
# Computing a Manhattan distance matrix.
mdistBirds <- dist(Birds, method = "manhattan")

# Hierachical clustering with average linkage.
clustBirds.avg.m <- hclust(mdistBirds, method = "average")

# Plot the plot!
ggdendrogram(clustBirds.avg.m, rotate = TRUE) + ggtitle("Average linkage, manhattan distance")
```

The dendrograms look quite the same! However, this might not always be
the case.

</details>

f)  Let's return to the research question: which birds have similar
    distribution in Norway during winter? How does your answer depend on
    linkage method and distance measure?

<details>

<summary><mark> Solution </mark></summary>

The answer depends on the distance measure and linkage method...

</details>

g)  You can read about other distance measures in the help file for
    distance measures in `R`. Write `?dist` in the console window and
    press enter. For instance, under **Details** you can read about
    "binary" distance. Use the code below to compute a binary distance
    matrix and plot the dendrogram with average linkage.

What do you think, does binary distance measure *abundance* or
*occurrence* distances between the birds?

```{r hclust4, eval = FALSE, echo = TRUE, indent = "   ", fig.align = "center"}
bdistBirds <- dist(Birds, method = "binary") # changing this to compute a binary distance matrix

clustBirds.avg.b <- hclust(bdistBirds, method = "average") %>%
   ggdendrogram(rotate = TRUE) + ggtitle("Binary distance & average linkage")

clustBirds.avg.b
```

<details>

<summary><mark> Solution </mark></summary>

It measures occurrence. </br> </br>

Under "Details" in `?dist`, the following is written about binary
distance:

"The vectors are regarded as binary bits, so non-zero elements are ‘on’
and zero elements are ‘off’. The distance is the proportion of bits in
which only one is on amongst those in which at least one is on".

We need only distinguish the zero-values from the non-zero in our
dataset. We will illustrate this with a simple example. Let's say we are
checking if two species of birds is present on 4 different sites.

Below there is one site where at least one is 'on', meaning at least one
bird species is present. For this site both are `on` (both species are
present at the same site). Therefore the distance is 0:

```{r}
foo <- rbind(c(0,1,0,0), c(0,1,0,0))
rownames(foo) <- c("Species A", "Species B")
colnames(foo) <- c("Site A", "Site B", "Site C", "Site D")
print(foo)
dist(foo, method = "binary")
```

Below there are two sites where at least one is `on`. In one of these,
only one is `on`. Therefore the distance is 0.5:

```{r}
foo <- rbind(c(1,0,0,0), c(1,1,0,0))
rownames(foo) <- c("Species A", "Species B")
colnames(foo) <- c("Site A", "Site B", "Site C", "Site D")
print(foo)
dist(foo, method = "binary")
```

If the two species cannot tolerate one another, we get a maximal
distance of 1:

```{r}
foo <- rbind(c(1,1,0,0), c(0,0,1,1))
rownames(foo) <- c("Species A", "Species B")
colnames(foo) <- c("Site A", "Site B", "Site C", "Site D")
print(foo)
dist(foo, method = "binary")
```

Even if we are counting the *number of birds* on each island, instead of
simply checking if it is there or not, the answer remains unchanged.
This is because we are comparing zero-values to non-zero-values:

```{r}
foo <- rbind(c(14,6,0,0), c(0,0,86,49))
rownames(foo) <- c("Species A", "Species B")
colnames(foo) <- c("Site A", "Site B", "Site C", "Site D")
print(foo)
dist(foo, method = "binary")
```

Let us return to the winterbirds data, and calculate the binary distance
between 4 of the bird species:

```{r}
dist(Birds4, method="binary")
```

Binary distance will always have a value of between 0 and 1, and is a
measure of occurrence between bird species. For instance, crow and
redpoll are the closest species in binary distance. This means that the
occurrence pattern of crow is more similar to the occurrence pattern of
common redpoll compared to any of the other 2 species.

</details>
