---
title: "Exercises STAT340 - Mixed Effects Models"
params:
  vis.forslag: false
output:
  html_document:
    toc: true
    theme: yeti
    code_folding: hide
    number_sections: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style>
/* Disable interaction with all details/summary */
details summary {
  pointer-events: none;   /* no mouse clicks */
  user-select: none;      /* cannot select text */
}

/* Optionally hide the disclosure triangle marker */
details summary::marker {
  content: "";
}
</style>
```

```{=html}
<style>
.main-container, .main-content { max-width: 100%; }
.info { background-color:aliceblue; }
.task { background-color:lightpink; }
</style>
```

```{r forslag, include = FALSE}
vis.forslag <- params$vis.forslag
```

```{r forklar forslag, echo = FALSE, results = "asis", include = vis.forslag}
cat('
<mark>
**LØSNINGSFORSLAG!**

Disse vises nedenfor på ulike måter:
<mark>Gulmerking</mark> av tekstforslag til løsning, eller kommentarer om alternative løsninger.
* Noe <button>Code</button> som var skjult i øvingsdokumentet.
<details>
<summary>
<mark>
Noen løsningsforslag er skjult inni hint.
</mark>
</summary>
Da må du utvide hintene for å se løsningsforslagene.  

</details>
')
```

<br> <br> <br>


You will need these libraries and datasets for these exercises:

```{r, echo = TRUE, message=FALSE, warning=FALSE}
library(BIAS.data)
library(nlme)
library(lme4)
library(ggplot2)
data(rats)
data(Orthodont)
data(cbpp)
```

# Ex-1: Random Intercept Model — `rats` data

The `rats` dataset from the `BIAS.data` package contains measurements of
`Glycogen` content in rat livers. Each rat (`Rat`) was measured across
three different liver sections (`Treatment`).

a.  Visualise the `Glycogen` content by `Rat` using a boxplot. What does
    the plot tell you about the variance *between* rats versus the
    variance *within* rats?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
boxplot(Glycogen ~ Rat, data = rats, col = "lightblue",
        ylab = "Glycogen Content", xlab = "Rat ID",
        main = "Glycogen by Rat")
```

The boxplot shows that different rats have systematically different
median glycogen values. Some (one of the two in our case) rats are
consistently high, others (the other one one of the two in our case)
consistently low. This means the **between-rat variance** is substantial
— measurements from the same rat are more similar to each other than
measurements from different rats. This is exactly the kind of dependency
that mixed models are designed to handle.

</details>

b.  Write down the mathematical formulation of a **random intercept
    model** for the `rats` data, explaining clearly what each term
    represents.

<details>

<summary><mark> Solution </mark></summary>

Let $y_{ij}$ be the Glycogen measurement for the $j$-th section of the
$i$-th rat. The random intercept model is:

$$y_{ij} = \beta_0 + b_i + \epsilon_{ij}$$

where:

-   $\beta_0$ is the global mean glycogen content (fixed intercept),
-   $b_i \sim N(0, \sigma^2_b)$ is the rat-specific random deviation
    from the global mean,
-   $\epsilon_{ij} \sim N(0, \sigma^2)$ is the residual measurement
    error.

The $b_i$ and $\epsilon_{ij}$ are assumed independent.

</details>

c.  Fit the random intercept model using `lme()` from the `nlme`
    package, with `Glycogen` as the response and `Rat` as the random
    effect grouping. Report the estimated variance components
    $\widehat\sigma_b$ and $\widehat\sigma$.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
rat_model <- lme(fixed = Glycogen ~ 1,
                 random = ~ 1 | Rat,
                 data = rats)
summary(rat_model)
```

From the output, the estimated standard deviations are approximately:

-   Between-rat (random intercept): $\widehat\sigma_b \approx 4.27$
-   Residual (within-rat): $\widehat\sigma \approx 9.26$

</details>

d.  Calculate the **Intraclass Correlation Coefficient (ICC)** using the
    variance components from part c. Interpret what the ICC means in
    this context.

<details>

<summary><mark> Solution </mark></summary>

The ICC formula is:

$$\text{ICC} = \frac{\sigma^2_b}{\sigma^2_b + \sigma^2}$$

Using the estimates above:

$$\sigma^2_b = 4.27^2 \approx 18.23, \quad \sigma^2 = 9.26^2 \approx 85.75$$

$$\text{ICC} = \frac{18.23}{18.23 + 85.75} \approx 0.175$$

```{r, echo=TRUE}
vc <- as.numeric(VarCorr(rat_model)[, "Variance"])
sigma2_b <- vc[1]
sigma2   <- vc[2]
icc <- sigma2_b / (sigma2_b + sigma2)
round(icc, 3)
```

**Interpretation:** About 17.5% of the total variation in Glycogen
content is explained purely by which rat the measurement came from. In
other words, two measurements from the **same** rat are expected to
correlate at $r \approx 0.175$, even before we account for treatment.

</details>

e.  Why would it be **wrong** to use a standard linear model (ignoring
    `Rat`) here? What problems would arise?

<details>

<summary><mark> Solution </mark></summary>

If we ignore the `Rat` grouping structure and fit
`lm(Glycogen ~ 1, data = rats)`, we assume all 36 observations are
independent. But measurements within the same rat are positively
correlated. This means:

1.  We **overestimate** the effective sample size (we have 6 rats, not
    36 independent units).
2.  Standard errors for the intercept will be **underestimated**.
3.  Any p-values will be **too small** — we inflate Type I error (false
    positives).
4.  The model's i.i.d. assumption is violated, so maximum likelihood
    inference is not valid.

</details>

f.  The `rats` dataset also records which liver `Treatment` section (1,
    2, or 3) each measurement came from. Extend the model by adding
    `Treatment` as a **nested random effect within `Rat`** — i.e., fit a
    model with a random intercept for `Rat` and an additional random
    intercept for `Treatment` nested within `Rat`. Write down the
    extended mathematical model and fit it in R. Does adding this nested
    level change the variance components substantially?

<details>

<summary><mark> Solution </mark></summary>

The extended nested model is:

$$y_{ijk} = \beta_0 + b_i + c_{j(i)} + \epsilon_{ijk}$$

where:

-   $b_i \sim N(0, \sigma^2_b)$: random intercept for rat $i$,
-   $c_{j(i)} \sim N(0, \sigma^2_c)$: random intercept for liver section
    $j$ **nested within** rat $i$ (each rat has its own section-level
    deviations),
-   $\epsilon_{ijk} \sim N(0, \sigma^2)$: residual error.

In `nlme`, nesting is expressed with the `/` operator in the `random`
argument:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
rat_nested <- lme(fixed = Glycogen ~ 1,
                  random = ~ 1 | Rat/Treatment,
                  data = rats)
summary(rat_nested)
```

The output now shows **three** variance components:

-   Between-rat variance ($\sigma^2_b$): how much rats differ on
    average.
-   Between-section-within-rat variance ($\sigma^2_c$): how much the
    three liver sections differ systematically within the same rat.
-   Residual variance ($\sigma^2$): pure measurement noise.

Comparing to the simpler model, $\widehat\sigma_c$ is small relative to
$\widehat\sigma_b$, so the treatment sections within a rat are not very
different from each other and the simpler model suffices. If
$\widehat\sigma_c$ was substantial, the nested structure would have been
important to capture.

</details>

------------------------------------------------------------------------

# Ex-2: Random Intercept + Random Slope — `fertilizer` data

The `fertilizer` dataset from `BIAS.data` tracks the root length of
individual plants over several weeks. Plants are assigned to either a
`fertilizer` or `control` treatment group.

a.  Plot the individual plant trajectories (root length over weeks)
    coloured by treatment group. Do plants appear to grow at different
    rates (slopes), or do they just start at different lengths
    (intercepts)?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE, message=FALSE, warning=FALSE}
data(fertilizer)
ggplot(fertilizer, aes(x = week, y = root, color = treatment)) +
  geom_line(aes(group = plant), alpha = 0.6) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Root length trajectories by plant",
       x = "Week", y = "Root length")
```

The plot reveals that plants differ in both their **starting length**
(intercepts vary) and their **rate of growth** (slopes vary). This
motivates a model with both random intercepts and random slopes.

</details>

b.  Write down the mathematical model that allows both the intercept and
    the slope over `week` to vary randomly by `plant`. Explain each
    component.

<details>

<summary><mark> Solution </mark></summary>

Let $y_{ij}$ be the root length of the $i$-th plant at week $x_{ij}$:

$$y_{ij} = (\beta_0 + b_{0i}) + (\beta_1 + b_{1i})\,x_{ij} + \epsilon_{ij}$$

where:

-   $\beta_0$: global average starting length (fixed intercept),
-   $\beta_1$: global average growth rate per week (fixed slope),
-   $b_{0i}$: plant-specific deviation from the average intercept,
-   $b_{1i}$: plant-specific deviation from the average slope,
-   $\epsilon_{ij} \sim N(0, \sigma^2)$: residual error.

The random effects $(b_{0i}, b_{1i})$ are assumed to follow a bivariate
normal distribution.

</details>

c.  Fit the random intercept + random slope model in R using `lme()`.
    Include `treatment` as a fixed effect and allow both the intercept
    and `week` slope to vary randomly by `plant`. Report the estimated
    fixed effects and interpret the `treatmentfertilizer` coefficient.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE, message=FALSE, warning=FALSE}
data(fertilizer)
res1 <- lme(fixed = root ~ treatment,
            random = ~ week | plant,
            data = fertilizer)
round(summary(res1)$tTable, 4)
```

**Interpretation:**

-   `(Intercept)` ≈ baseline root length for the control group.
-   `treatmentfertilizer` ≈ 1.04: On average, plants receiving
    fertilizer are **1.04 units longer** than control plants. This
    effect is highly significant ($p = 0.0005$).
-   Note that `week` is modelled as a **random slope**, so there is no
    single fixed-effect estimate for it; each plant gets its own slope.

</details>

d.  Examine the residual diagnostic plots (fitted vs residuals and QQ
    plot of standardised residuals). Do the model assumptions appear to
    be satisfied?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE, fig.height=3.5, fig.width=7}
par(mfrow = c(1, 2))
plot(fitted(res1), resid(res1, type = "p"),
     main = "Fitted vs Residuals", ylab = "Std. Residuals", xlab = "Fitted")
abline(h = 0, lty = 2)
qqnorm(resid(res1, type = "p"))
qqline(resid(res1, type = "p"))
par(mfrow = c(1, 1))
```

The residuals appear randomly scattered around zero with no clear
pattern, and the QQ plot follows the diagonal line well. The model
assumptions of homoscedasticity and normality seem reasonably satisfied.

</details>

e.  Calculate the ICC **at week 0** using the random effects standard
    deviations from the model summary. Interpret the result.

<details>

<summary><mark> Solution </mark></summary>

From the `summary(res1)` output:

-   $\widehat\sigma_{b0} \approx 2.86$ (random intercept SD)
-   $\widehat\sigma \approx 0.50$ (residual SD)

At week 0, the between-plant variance equals only the intercept
variance:

$$\text{ICC}_{week=0} = \frac{\sigma^2_{b0}}{\sigma^2_{b0} + \sigma^2} = \frac{2.86^2}{2.86^2 + 0.50^2} = \frac{8.18}{8.18 + 0.25} \approx 0.97$$

```{r, echo=TRUE}
sigma_b0 <- as.numeric(VarCorr(res1)["(Intercept)", "StdDev"])
sigma_res <- as.numeric(VarCorr(res1)["Residual",   "StdDev"])
icc_week0 <- sigma_b0^2 / (sigma_b0^2 + sigma_res^2)
round(icc_week0, 3)
```

**Interpretation:** At the start of the experiment (week 0), almost
**97% of the variation** in root length is due to which plant we are
measuring — i.e., the plant's individual genetics/seed quality. As time
progresses and the random slopes diverge, this proportion changes.

</details>

f.  Using the model from part c. (`res1`, with `treatment` as a fixed
    effect and `week` as a random slope by `plant`), produce predictions
    at two levels:

    (i) **Population-level** (`level = 0`): the average expected root
        length for a **control** plant and a **fertilizer** plant over
        weeks 1–5, using only the fixed effects.
    (ii) **Individual-level** (`level = 1`): the predicted root length
         for two specific plants from the dataset over the same weeks,
         using their estimated random effects.

    Comment on the difference between the two types of predictions.

<details>

<summary><mark> Solution </mark></summary>

We refit `res1` as in part c. and use `predict()` with `level = 0`
(population) and `level = 1` (individual).

```{r, echo=TRUE, message=FALSE, warning=FALSE}
data(fertilizer)
res1 <- lme(fixed = root ~ treatment,
            random = ~ week | plant,
            data = fertilizer)

# (i) Population-level: same prediction for any plant of that treatment
new_pop <- data.frame(
  plant     = rep(unique(fertilizer$plant)[1], 10), # id needed but ignored
  week      = rep(1:5, 2),
  treatment = rep(c("control", "fertilizer"), each = 5)
)
new_pop$pred_pop <- predict(res1, newdata = new_pop, level = 0)
print(new_pop)
```

```{r, echo=TRUE}
# (ii) Individual-level: predictions for two specific plants
plants_used <- unique(fertilizer$plant)[c(1, 5)]
new_ind <- data.frame(
  plant     = rep(plants_used, each = 5),
  week      = rep(1:5, 2),
  treatment = rep(
    fertilizer$treatment[match(plants_used, fertilizer$plant)],
    each = 5
  )
)
new_ind$pred_ind <- predict(res1, newdata = new_ind, level = 1)
print(new_ind)
```

**Key difference:**

-   **Level 0 (population)** uses only the fixed effect
    $\widehat\beta_0 + \widehat\beta_\text{treatment}$. The prediction
    is identical for every plant of that treatment group — it is the
    model's best guess for a **new, unseen** plant.
-   **Level 1 (individual)** adds the plant's estimated random intercept
    $\widehat b_{0i}$ and random slope $\widehat b_{1i}$, giving the
    best prediction *for that specific plant* based on its observed
    data. Individual predictions vary across plants and are closer to
    the actual observed values.

</details>

------------------------------------------------------------------------

# Ex-3: Hierarchical Nested Design — `Orthodont` data

The `Orthodont` dataset (available via `nlme`) tracks the pituitary
`distance` (mm) in children at ages 8, 10, 12, and 14. Children
(`Subject`) are nested within `Sex`.

a.  Produce a spaghetti plot of `distance` over `age`, coloured by
    `Sex`, with one line per subject. Describe the key features you
    observe.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE, message=FALSE, fig.height=4.5, fig.width=7}
data(Orthodont)
ggplot(Orthodont, aes(x = age, y = distance, color = Sex)) +
  geom_line(aes(group = Subject), alpha = 0.6) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Pituitary distance by age and sex",
       x = "Age (years)", y = "Distance (mm)")
```

Key observations:

-   Both boys and girls show **increasing distance** with age.
-   Males tend to have **higher** baseline distances than females.
-   There is substantial **between-child variability** in both starting
    distance and rate of growth.

</details>

b.  Fit a **random intercept** model with `age` and `Sex` as fixed
    effects and `Subject` as the random grouping factor. Write the
    mathematical model first, then fit it in R using `lme()`.

<details>

<summary><mark> Solution </mark></summary>

Let $y_{ijk}$ be the distance for subject $i$ of sex $k$ at age
$x_{ijk}$:

$$y_{ijk} = \beta_0 + \beta_1 \text{Age}_{ijk} + \beta_2 \text{Sex}_k + b_{i(k)} + \epsilon_{ijk}$$

where $b_{i(k)} \sim N(0, \sigma^2_b)$ is the random intercept for
subject $i$ nested within sex $k$.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
res3 <- lme(fixed = distance ~ age + Sex,
            random = ~ 1 | Subject,
            data = Orthodont)
summary(res3)
```

</details>

c.  Interpret the fixed effect coefficients for `age` and `SexFemale`.
    Are both significant?

<details>

<summary><mark> Solution </mark></summary>

From the fixed effects table:

-   **`age`** ≈ 0.66: For each additional year of age, the pituitary
    distance increases by about **0.66 mm** on average. This is highly
    significant ($p < 0.001$).

-   **`SexFemale`** ≈ −2.32: Females have a baseline distance that is
    about **2.32 mm lower** than males. This is also highly significant
    ($p < 0.001$).

Because this is a **random intercept** model, the growth rate of 0.66
mm/year is assumed to be the **same for all children** — only their
baseline distances differ.

</details>

d.  Calculate the ICC for the `Orthodont` random intercept model and
    interpret it.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
vc3 <- as.numeric(VarCorr(res3)[, "Variance"])
sigma2_b3  <- vc3[1]
sigma2_res3 <- vc3[2]
icc3 <- sigma2_b3 / (sigma2_b3 + sigma2_res3)
round(icc3, 3)
```

Using $\widehat\sigma_b \approx 1.807$ and
$\widehat\sigma \approx 1.432$:

$$\text{ICC} = \frac{1.807^2}{1.807^2 + 1.432^2} = \frac{3.265}{3.265 + 2.051} \approx 0.614$$

**Interpretation:** About **61.4%** of the remaining variation in
pituitary distance (after accounting for age and sex) is due to
permanent biological differences between individual children. The four
measurements within the same child are highly correlated.

</details>

e.  Now extend the model by adding a **random slope** for `age` (in
    addition to the random intercept) by `Subject`. Fit this extended
    model and compare the random effects standard deviations to the
    random-intercept-only model. What does adding a random slope
    capture?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE, message=FALSE, warning=FALSE}
res4 <- lme(fixed = distance ~ age + Sex,
            random = ~ age | Subject,
            data = Orthodont)
summary(res4)
```

The extended model now estimates:

-   A standard deviation for the **random intercept** (between-child
    baseline differences).
-   A standard deviation for the **random slope on age** (between-child
    differences in growth rate).

Adding a random slope for `age` allows each child to have their own
individual growth trajectory — some children grow faster, some slower —
rather than forcing everyone to grow at the global average rate of
\~0.66 mm/year.

</details>

f.  Compare the random-intercept-only model (`res3`) and the
    random-intercept + random-slope model (`res4`) using **AIC** and
    **BIC**. Refit both models with **ML** (not REML) before comparing,
    since the two models differ only in their random effects structure.
    Which model does each criterion prefer, and does the conclusion
    differ between AIC and BIC? Explain why BIC tends to be more
    conservative (prefers simpler models) than AIC.

<details>

<summary><mark> Solution </mark></summary>

When comparing models with **different random effects**, we must use ML
(not REML), because REML likelihoods are not comparable across models
with different random-effects structures.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
res3_ml <- lme(fixed = distance ~ age + Sex,
               random = ~ 1 | Subject,
               data = Orthodont,
               method = "ML")
res4_ml <- lme(fixed = distance ~ age + Sex,
               random = ~ age | Subject,
               data = Orthodont,
               method = "ML",
               control = lmeControl(maxIter = 200, msMaxIter = 200))

data.frame(
  Model = c("Random intercept only", "Random intercept + slope"),
  df    = c(attr(logLik(res3_ml), "df"), attr(logLik(res4_ml), "df")),
  AIC   = round(c(AIC(res3_ml), AIC(res4_ml)), 2),
  BIC   = round(c(BIC(res3_ml), BIC(res4_ml)), 2)
)
```

**Conclusion:**

Simpler model is preferred in our cases w.r.t. to both criteria.

</details>

------------------------------------------------------------------------

# Ex-4: Generalized Linear Mixed Model — `cbpp` data

The `cbpp` dataset (from `lme4`) records cases of Contagious Bovine
Pleuropneumonia (CBPP) across several herds of cattle over four time
periods.

Variables:

-   `incidence`: Number of sick cows in the herd (successes).
-   `size`: Total number of cows in the herd (trials).
-   `period`: Time period (1–4), treated as a fixed effect.
-   `herd`: The specific herd (random effect — some herds may be in more
    disease-prone environments).

a.  Why is it important to include `herd` as a **random** rather than a
    **fixed** effect in this model?

<details>

<summary><mark> Solution </mark></summary>

The herds in the study are a **random sample** from a larger population
of herds. We are not interested in the specific disease rates of these
exact 15 herds, but in the general herd-to-herd variation. Treating
`herd` as a random effect allows us to:

1.  Properly account for the correlation among measurements within the
    same herd across periods.
2.  Estimate the **variance component** due to herd-level factors
    (environment, management, genetics).
3.  Generalise conclusions beyond the specific herds observed.

If we treated `herd` as fixed, we would consume many degrees of freedom
and only draw conclusions about those specific 15 herds.

</details>

b.  Write the GLMM formulation for this model. What link function is
    used, and why?

<details>

<summary><mark> Solution </mark></summary>

Let $y_{ij}$ be the incidence count in herd $i$ during period $j$, with
$n_{ij}$ total cows. We model:

$$y_{ij} \sim \text{Binomial}(n_{ij},\, \pi_{ij})$$

$$\text{logit}(\pi_{ij}) = \beta_0 + \beta_1 \text{Period}_{ij} + b_i, \quad b_i \sim N(0, \sigma^2_b)$$

The **logit link** is used because the response is a proportion (disease
incidence rate), and the logit maps probabilities in $(0,1)$ to the
whole real line, making it compatible with the linear predictor.

</details>

c.  Fit the GLMM using `glmer()` from `lme4` with `period` as a fixed
    effect and a random intercept for `herd`. Use `family = binomial`.
    Report the fixed effect estimates.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE, warning=FALSE, message=FALSE}
glmm_mod <- glmer(cbind(incidence, size - incidence) ~
                    period + (1 | herd),
                  family = binomial(link = "logit"),
                  data = cbpp)
round(summary(glmm_mod)$coefficients, 4)
```

The estimated fixed effects are on the **log-odds scale**:

-   `(Intercept)`: The log-odds of disease incidence in the reference
    period (period 1) for an average herd.
-   `period2`, `period3`, `period4`: The change in log-odds relative to
    period 1. All are negative, indicating the disease rate is
    **decreasing over the four periods**.

</details>

d.  Extract the estimated random intercept standard deviation
    $\widehat\sigma_b$ from the fitted GLMM. What does it tell us about
    herd-to-herd variability in disease risk?

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE}
sigma_b_glmm <- as.data.frame(VarCorr(glmm_mod))$sdcor[1]
round(sigma_b_glmm, 3)
```

The estimated $\widehat\sigma_b$ gives the standard deviation of the
herd-level random intercepts **on the log-odds scale**. A value
$\widehat\sigma_b \approx 0.6$–$1.0$ (typical for these data) means that
herds differ noticeably in their baseline disease susceptibility. Herds
with positive $b_i$ have higher log-odds (higher incidence), while those
with negative $b_i$ are more resistant. Ignoring this variation would
lead to incorrect (too narrow) confidence intervals for the period
effects.

</details>

f.  Suppose we **did not know the herd sizes** (`size`) and only
    observed the raw incidence counts. Fit a **Poisson GLMM** with
    `period` as a fixed effect and a random intercept for `herd`.
    Explain when a Poisson model is an appropriate alternative to the
    Binomial model for count data, and what the key difference in
    interpretation is.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE, warning=FALSE, message=FALSE}
glmm_pois <- glmer(incidence ~ period + (1 | herd),
                   family = poisson(link = "log"),
                   data = cbpp)
round(summary(glmm_pois)$coefficients, 4)
```

**When is the Poisson model appropriate?**

The Binomial model requires knowing the **total number of trials**
(`size`) so it can model the *proportion* that are successes. The
Poisson model is appropriate when:

1.  We only observe **raw counts** of events without knowing the
    denominator (total at risk).
2.  The counts arise from a process where events are rare and
    approximately independent (e.g., number of disease cases per time
    period in an unlimited population).
3.  The expected count is well below the maximum possible count, so the
    Binomial approximates a Poisson.

**Difference in interpretation:**

-   **Binomial GLMM**: Fixed effects are on the **log-odds** scale; we
    model the *probability* of an individual cow being sick given herd
    size.
-   **Poisson GLMM**: Fixed effects are on the **log-count** scale; we
    model the *expected number* of cases per period, ignoring herd size.
    A one-unit change in a predictor multiplies the expected count by
    $e^{\widehat\beta}$.

In practice, because we *do* know `size` here, the Binomial model is
more appropriate — it uses all available information. The Poisson model
in this context discards the denominator and effectively assumes all
herds have the same exposure, which is not true.

</details>

------------------------------------------------------------------------

# Ex-5: Temporal Autocorrelation — `Orthodont` with CAR(1)

Instead of treating repeated measurements on children as exchangeable
(standard random intercept), we can explicitly model the fact that
measurements closer in time may be more correlated than distant ones.

a.  Fit a model with fixed effects `age` and `Sex`, a random intercept
    for `Subject`, **and** a continuous-time AR(1) (`corCAR1`)
    correlation structure across `age`. Report the estimated
    autocorrelation parameter $\Phi$.

<details>

<summary><mark> Solution </mark></summary>

```{r, echo=TRUE, message=FALSE, warning=FALSE}
model_car1 <- lme(fixed = distance ~ age + Sex,
                  random = ~ 1 | Subject,
                  correlation = corCAR1(form = ~ 1 + age | Subject),
                  data = Orthodont)
summary(model_car1)$modelStruct$corStruct
```

The estimated $\widehat\Phi \approx 0.24$ represents the temporal
autocorrelation between consecutive age measurements for the same child.
A value of 0.24 indicates a **moderate positive autocorrelation**: a
child who is above the expected distance at age 8 tends also to be above
expected at age 10, but this carry-over decays over time.

</details>

b.  *(Optional — hard)* What does a CAR(1) correlation structure assume
    about how the correlation between two measurements changes as the
    time gap between them increases?

<details>

<summary><mark> Solution </mark></summary>

A continuous-time AR(1) (CAR1) structure assumes that the correlation
between two measurements on the same subject at times $t_1$ and $t_2$
decays **exponentially** with the time difference:

$$\text{Corr}(y_{t_1}, y_{t_2}) = \Phi^{|t_1 - t_2|}$$

So:

-   Measurements 2 years apart (e.g., age 8 and 10) correlate as
    $\Phi^2$.
-   Measurements 4 years apart (e.g., age 8 and 12) correlate as
    $\Phi^4$.
-   Measurements 6 years apart (e.g., age 8 and 14) correlate as
    $\Phi^6$.

As $|t_1 - t_2| \to \infty$, the correlation $\to 0$. This is more
realistic than assuming all repeated measurements correlate equally
(compound symmetry / random intercept only).

</details>

c.  Why is it important to use **REML** (the default in `lme()`) rather
    than standard **ML** when estimating variance components in linear
    mixed models?

<details>

<summary><mark> Solution </mark></summary>

When we estimate variance components using standard Maximum Likelihood
(ML), the estimator is **biased downward** — it is too small. This
happens because ML does not account for the degrees of freedom used up
by estimating the fixed effects.

**REML (Restricted Maximum Likelihood)** corrects for this by first
"removing" the fixed effects from the likelihood (working with the
residuals rather than the raw data), and then estimating the variance
components. The analogy is the divisor $n-1$ in the sample variance
formula: we adjust for having estimated the mean.

Key rules:

-   Use **REML** (default) when your goal is accurate variance component
    estimates.
-   Use **ML** when you need to compare models with **different fixed
    effects** via likelihood ratio tests, since REML likelihoods are not
    comparable across different fixed-effect structures.

</details>
